---
title: "Glycosylation traits"
author: "Copyright 2016-2023 Karli Reiding, Viktoria Dotz, Gerda Vreeker, Guinevere Lageveen-Kammeijer, Marco Bladergroen"
date: "version: 1.2.3, 2025-03-06, executed `r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
editor_options:
  chunk_output_type: inline
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding,
                                                         output_file = file.path(dirname(inputFile), 
                                                         paste0('Glycosylation Traits ',
                                                         format(Sys.time(), '%Y-%m-%d %H_%M_%S'), '.html'))) })
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Version history:

Version 0.94 (Karli Reiding, Viktoria Dotz, Gerda Vreeker): 2019-03-27
<ul>
Last modified by Viktoria Dotz who found mistakes in the calculation of A4FG and A4F0G\
Traits that Gerda Vreeker created in 2018 added.
</ul>
Version 0.94_GSM (Guinevere Lageveen-Kammeijer): 2020-05-06
<ul>
Guinevere added paucimannosidic trait (H3N2), unknown Classification (H3N3F0/1, H3H4F0/1/2), trait ALN and changed the high mannose trait (F=0)
</ul>
Versions 0.95/0.96 (Marco Bladergroen): 2021-04-29
<ul>
Some calculations simplified, high-mannose (MM): H4N2 added\
Filenames and settings moved to top of script, excel possibility added.\
Sialylation dependent on galactosylation extended to 3- and 4-antennary structures\
Some traits for Visceral Leishmaniasis cohort (Gabriane Porcino) added.
</ul>
Version 0.97 (Marco Bladergroen): 2021-06-04
<ul>Derived_traits_curated added as extra sheet. Columns with all 1, all 0 or all NA removed from derived list</ul>
Version 0.98 (Marco Bladergroen): 2021-07-01
<ul>Sheet with all traits combined (direct and derived traits) added to excel output</ul>
Version 1.0  (Marco Bladergroen): 2022-08-19
<ul>Format changed from r-script to R-markdown.</ul>
Version 1.1  (Marco Bladergroen): 2024-04-11
<ul>Script adapted to mouse glycans (including N-glycolyl neuraminic acid with modifications).</ul>
Version 1.2.3 (Juan-Carlos Rojas Echeverri): 2025-03-06
<ul>Missing building blocks added. Calculations moved to beginning of script. (Among other things)</ul>

## Used packages
```{r packages, message=FALSE, warning=FALSE, include=TRUE}
library(openxlsx)
library(stringr)
library(tidyverse)
```

## Global variables/settings

```{r settings, include=TRUE}
wd <- getwd()
import_file_name <- "<Filename of your curated data>.xlsx"
sheetName <- "renormalized"
export_file_name <- paste("<Your output filename>", format(Sys.time(), "%Y-%m-%d %H_%M_%S"), ".xlsx", sep="")
fname_data = paste(wd,import_file_name,sep="/")
fname_export = paste(wd,export_file_name,sep="/")

taxonomy <- "human" # other taxonomy possible is mouse/murine
```

## Requested calculations

```{r requested_calculations, include=TRUE}
requested_calculations <- matrix("",0,1)

  # Overall glycan types
  requested_calculations["T"] = "T"
  requested_calculations["TP"] = "TP / T"
  requested_calculations["TU"] = "TU / T"
  requested_calculations["TM"] = "TM / T"
  requested_calculations["THy"] = "THy / T"
  requested_calculations["TC"] = "TC / T"
  requested_calculations["TAc"] = "TAc / T"
  # requested_calculations["TG"] = "TG / T"
  # requested_calculations["TS"] = "TS / T"
  # requested_calculations["MHy"] = "TM / THy"
  # requested_calculations["TE"] = "TE / TS"
  # requested_calculations["TL"] = "TL / TS"
  
  # High mannose size
  requested_calculations["MM"] = "( 4 * TM4 + 5 * TM5 + 6 * TM6 + 7 * TM7 + 8 * TM8 + 9 * TM9 + 10 * TM10 + 11 * TM11 + 12 * TM12 ) / TM"
  
  # Complex type distribution
  requested_calculations["CA1"] = "TA1 / TC"
  requested_calculations["CA2"] = "TA2 / TC"
  requested_calculations["CA3"] = "TA3 / TC"
  requested_calculations["CA4"] = "TA4 / TC"
  #requested_calculations["CF0"] = "TF0 / TC"
  requested_calculations["CF"] = "TF / TC"
  #requested_calculations["CFa"] = "TFa / TC"
  #requested_calculations["CB0"] = "TB0 / TC"
  requested_calculations["CB"] = "TB / TC"
  requested_calculations["CG"] = "TG / TC"
  requested_calculations["CS"] = "TS / TC"
  #requested_calculations["CE"] = "TE / TC"
  #requested_calculations["CL"] = "TL / TC"
  
  # Underived sums
  # requested_calculations["TA2FS0"] = "TA2FS0"
  # requested_calculations["TA2FS"] = "TA2FS"
  # requested_calculations["TA2FG0S0"] = "TA2FG0S0"
  # requested_calculations["TA2FG1S0"] = "TA2FG1S0"
  # requested_calculations["TA2FG2S0"] = "TA2FG2S0"
  # requested_calculations["TA2L0"] = "TA2L0"
  # requested_calculations["TA3L0"] = "TA3L0"
  # requested_calculations["TA4L0"] = "TA4L0"
  # requested_calculations["TA1L"] = "TA1L"
  # requested_calculations["TA2L"] = "TA2L"
  # requested_calculations["TA3L"] = "TA3L"
  # requested_calculations["TA4L"] = "TA4L"
  # requested_calculations["TA1E"] = "TA1E"
  # requested_calculations["TA2E"] = "TA2E"
  # requested_calculations["TA3E"] = "TA3E"
  # requested_calculations["TA4E"] = "TA4E"
  requested_calculations["TA2S"] = "TA2S"
  requested_calculations["TA3S"] = "TA3S"
  requested_calculations["TA4S"] = "TA4S"
  
  #Antennaerity
  #requested_calculations["A1"] = "A1"
  #requested_calculations["A2"] = "A2"
  #requested_calculations["A3"] = "A3"
  #requested_calculations["A4"] = "A4"
  #requested_calculations["TA0"] = "( TP + TM ) / T"
  #requested_calculations["TA1"] = "TA1"
  requested_calculations["TA1"] = "A1 / T"
  requested_calculations["TA2"] = "A2 / T"
  requested_calculations["TA3"] = "A3 / T"
  requested_calculations["TA4"] = "A4 / T"
  #requested_calculations["TALN"] = "ALN / T"
  #requested_calculations["CA1"] = "( A1 - TU - H5N3 - H5N3F1 - H5N3F1S1 - H5N3S1 - H6N3F1 - H6N3F1S1 - H6N3S1 ) / TC"
  #requested_calculations["CA2"] = "( A2 - H6N4 ) / TC"
  #requested_calculations["CA3"] = "A3 / TC"
  #requested_calculations["CA4"] = "A4 / TC"
  #requested_calculations["CALN"] = "ALN / TC"
  
  
  # Fucosylation
  # requested_calculations["TF0"] = "( TM + H3N2 + H3N3 + H4N3 + H5N3 + H5N3S1 + H6N3S1 + H6N4 + TF0 ) / T"
  # requested_calculations["TF"] = "( H3N2F1 + H3N3F1 + H4N3F1 + H4N3F2 + H5N3F1 + H5N3F1S1 + H6N3F1 + H6N3F1S1 + TF ) / T"
  requested_calculations["TF"] = "TF / T"
  # requested_calculations["TF_1"] = "( H3N3F1 + H4N3F1 + H5N3F1 + TF ) / T"
  # requested_calculations["TFa"] = "( TFa + H4N3F2 ) / T"
  # requested_calculations["A1F0"] = "TA1F0 / TA1"
  # requested_calculations["A2F0"] = "TA2F0 / TA2"
  # requested_calculations["A3F0"] = "TA3F0 / TA3"
  # requested_calculations["A4F0"] = "TA4F0 / TA4"
  requested_calculations["A1F"] = "TA1F / TA1"
  requested_calculations["A2F"] = "TA2F / TA2"
  requested_calculations["A3F"] = "TA3F / TA3"
  requested_calculations["A4F"] = "TA4F / TA4"
  requested_calculations["A1Fa"] = "TA1Fa / TA1"
  requested_calculations["A2Fa"] = "TA2Fa / TA2"
  requested_calculations["A3Fa"] = "TA3Fa / TA3"
  requested_calculations["A4Fa"] = "TA4Fa / TA4"
  
  # Fucosylation of (non)sialylated species
  # requested_calculations["A1S0F"] = "TA1FS0 / TA1S0"
  requested_calculations["A2S0F"] = "TA2FS0 / TA2S0"
  requested_calculations["A3S0F"] = "TA3FS0 / TA3S0"
  requested_calculations["A4S0F"] = "TA4FS0 / TA4S0"
  # requested_calculations["A1L0F"] = "TA1FL0 / TA1L0"
  # requested_calculations["A2L0F"] = "TA2FL0 / TA2L0"
  # requested_calculations["A3L0F"] = "TA3FL0 / TA3L0"
  # requested_calculations["A4L0F"] = "TA4FL0 / TA4L0"
  # requested_calculations["A1E0F"] = "TA1FE0 / TA1E0"
  # requested_calculations["A2E0F"] = "TA2FE0 / TA2E0"
  # requested_calculations["A3E0F"] = "TA3FE0 / TA3E0"
  # requested_calculations["A4E0F"] = "TA4FE0 / TA4E0"
  # requested_calculations["A1SF"] = "TA1FS / TA1S"
  requested_calculations["A2SF"] = "TA2FS / TA2S"
  requested_calculations["A3SF"] = "TA3FS / TA3S"
  requested_calculations["A4SF"] = "TA4FS / TA4S"
  # requested_calculations["A1LF"] = "TA1FL / TA1L"
  # requested_calculations["A2LF"] = "TA2FL / TA2L"
  # requested_calculations["A3LF"] = "TA3FL / TA3L"
  # requested_calculations["A4LF"] = "TA4FL / TA4L"
  # requested_calculations["A1EF"] = "TA1FE / TA1E"
  # requested_calculations["A2EF"] = "TA2FE / TA2E"
  # requested_calculations["A3EF"] = "TA3FE / TA3E"
  # requested_calculations["A4EF"] = "TA4FE / TA4E"
  
  # Bisection
  requested_calculations["TB"] = "TA2B / T"
  requested_calculations["A2B"] = "TA2B / TA2"
  requested_calculations["A2F0B"] = "TA2F0B / TA2F0"
  requested_calculations["A2FB"] = "TA2FB / TA2F"
  requested_calculations["A2S0B"] = "TA2S0B / TA2S0"
  requested_calculations["A2SB"] = "TA2SB / TA2S"
  requested_calculations["A2F0S0B"] = "TA2F0S0B / TA2F0S0"
requested_calculations["A2F0SB"] = "TA2F0SB / TA2F0S"
  requested_calculations["A2FS0B"] = "TA2FS0B / TA2FS0"
  requested_calculations["A2FSB"] = "TA2FSB / TA2FS"
  
  # Galactosylation
  requested_calculations["A1G"] = "( 0/1 * TA1G0 + 1/1 * TA1G1 ) / TA1"
  requested_calculations["A2G"] = "( 0/2 * TA2G0 + 1/2 * TA2G1 + 2/2 * TA2G2 ) / TA2"
  requested_calculations["A3G"] = "( 0/3 * TA3G0 + 1/3 * TA3G1 + 2/3 * TA3G2 + 3/3 * TA3G3 ) / TA3"
  requested_calculations["A4G"] = "( 0/4 * TA4G0 + 1/4 * TA4G1 + 2/4 * TA4G2 + 3/4 * TA4G3 + 4/4 * TA4G4 ) / TA4"
  # requested_calculations["A1F0G"] = "( 0/1 * TA1F0G0 + 1/1 * TA1F0G1 ) / TA1F0"
  # requested_calculations["A2F0G"] = "( 0/2 * TA2F0G0 + 1/2 * TA2F0G1 + 2/2 * TA2F0G2 ) / TA2F0"
  # requested_calculations["A3F0G"] = "( 0/3 * TA3F0G0 + 1/3 * TA3F0G1 + 2/3 * TA3F0G2 + 3/3 * TA3F0G3 ) / TA3F0"
  # requested_calculations["A4F0G"] = "( 0/4 * TA4F0G0 + 1/4 * TA4F0G1 + 2/4 * TA4F0G2 + 3/4 * TA4F0G3 + 4/4 * TA4F0G4 ) / TA4F0"
  # requested_calculations["A1FG"] = "( 0/1 * TA1FG0 + 1/1 * TA1FG1 ) / TA1F"
  # requested_calculations["A2FG"] = "( 0/2 * TA2FG0 + 1/2 * TA2FG1 + 2/2 * TA2FG2 ) / TA2F"
  # requested_calculations["A3FG"] = "( 0/3 * TA3FG0 + 1/3 * TA3FG1 + 2/3 * TA3FG2 + 3/3 * TA3FG3 ) / TA3F"
  # requested_calculations["A4FG"] = "( 0/4 * TA4FG0 + 1/4 * TA4FG1 + 2/4 * TA4FG2 + 3/4 * TA4FG3 + 4/4 * TA4FG4 ) / TA4F"
  
  # Sialylation dependent on galactosylation
  # requested_calculations["A2S0G"] = "( 0/2 * TA2G0S0 + 1/2 * TA2G1S0 + 2/2 * TA2G2S0 ) / TA2S0"
  requested_calculations["A2SG"] = "( 0/2 * TA2G0S + 1/2 * TA2G1S + 2/2 * TA2G2S ) / TA2S"   
  requested_calculations["A2F0S0G"] = "( 0/2 * TA2F0G0S0 + 1/2 * TA2F0G1S0 + 2/2 * TA2F0G2S0 ) / TA2F0S0"
  requested_calculations["A2FS0G"] = "( 0/2 * TA2FG0S0 + 1/2 * TA2FG1S0 + 2/2 * TA2FG2S0 ) / TA2FS0"
  requested_calculations["A2F0SG"] = "( 0/2 * TA2F0G0S + 1/2 * TA2F0G1S + 2/2 * TA2F0G2S ) / TA2F0S"
  requested_calculations["A2FSG"] = "( 0/2 * TA2FG0S + 1/2 * TA2FG1S + 2/2 * TA2FG2S ) / TA2FS"
  requested_calculations["A3S0G"] = "( 0/3 * TA3G0S0 + 1/3 * TA3G1S0 + 2/3 * TA3G2S0 + 3/3 * TA3G3S0 ) / TA3S0"
  requested_calculations["A3SG"] = "( 0/3 * TA3G0S + 1/3 * TA3G1S + 2/3 * TA3G2S + 3/3 * TA3G3S ) / TA3S"   
  requested_calculations["A3F0S0G"] = "( 0/3 * TA3F0G0S0 + 1/3 * TA3F0G1S0 + 2/3 * TA3F0G2S0 + 3/3 * TA3F0G3S0) / TA3F0S0"
  requested_calculations["A3FS0G"] = "( 0/3 * TA3FG0S0 + 1/3 * TA3FG1S0 + 2/3 * TA2FG2S0 + 3/3 * TA3FG3S0 ) / TA3FS0"
  requested_calculations["A3F0SG"] = "( 0/3 * TA3F0G0S + 1/3 * TA3F0G1S + 2/3 * TA3F0G2S + 3/3 * TA3F0G3S ) / TA3F0S"
  requested_calculations["A3FSG"] = "( 0/3 * TA3FG0S + 1/3 * TA3FG1S + 2/3 * TA3FG2S + 3/3 * TA3FG3S ) / TA3FS"
  requested_calculations["A4S0G"] = "( 0/4 * TA4G0S0 + 1/4 * TA4G1S0 + 2/4 * TA4G2S0 + 3/4 * TA4G3S0 + 4/4 * TA4G4S0  ) / TA4S0"
  requested_calculations["A4SG"] = "( 0/4 * TA4G0S + 1/4 * TA4G1S + 2/4 * TA4G2S + 3/4 * TA4G3S + 4/4 * TA4G4S ) / TA4S"   
  requested_calculations["A4F0S0G"] = "( 0/4 * TA4F0G0S0 + 1/4 * TA4F0G1S0 + 2/4 * TA4F0G2S0 + 3/4 * TA4F0G3S0 + 4/4 * TA4F0G4S0 ) / TA4F0S0"
  requested_calculations["A4FS0G"] = "( 0/4 * TA4FG0S0 + 1/4 * TA4FG1S0 + 2/4 * TA4FG2S0 + 3/4 * TA4FG3S0 + 4/4 * TA4FG4S0 ) / TA4FS0"
  requested_calculations["A4F0SG"] = "( 0/4 * TA4F0G0S + 1/4 * TA4F0G1S + 2/4 * TA4F0G2S + 3/4 * TA4F0G3S + 4/4 * TA4F0G4S ) / TA4F0S"
  requested_calculations["A4FSG"] = "( 0/4 * TA4FG0S + 1/4 * TA4FG1S + 2/4 * TA4FG2S + 3/4 * TA4FG3S + 4/4 * TA4FG4S ) / TA4FS"
  
  # Sialylation
  #requested_calculations["A1S"] = "( 0/1 * TA1S0 + 1/1 * TA1S1 ) / TA1"
  #requested_calculations["TA1S"] = "TA1S1 / T"
  #requested_calculations["A2S"] = "( 0/2 * TA2S0 + 1/2 * TA2S1 + 2/2 * TA2S2 ) / TA2"
  #requested_calculations["TA2S"] = "( TA2S1 + TA2S2 ) / T"
  #requested_calculations["A3S"] = "( 0/3 * TA3S0 + 1/3 * TA3S1 + 2/3 * TA3S2 + 3/3 * TA3S3 ) / TA3"
  #requested_calculations["TA3S"] = "( TA3S1 + TA3S2 + TA3S3 ) / T"
  #requested_calculations["A4S"] = "( 0/4 * TA4S0 + 1/4 * TA4S1 + 2/4 * TA4S2 + 3/4 * TA4S3 + 4/4 * TA4S4 ) / TA4"
  #requested_calculations["TA4S"] = "( TA4S1 + TA4S2 + TA4S3 + TA4S4 ) / T"
  # requested_calculations["A1F0S"] = "( 0/1 * TA1F0S0 + 1/1 * TA1F0S1 ) / TA1F0"
  requested_calculations["A2F0S"] = "( 0/2 * TA2F0S0 + 1/2 * TA2F0S1 + 2/2 * TA2F0S2 ) / TA2F0"
  requested_calculations["A3F0S"] = "( 0/3 * TA3F0S0 + 1/3 * TA3F0S1 + 2/3 * TA3F0S2 + 3/3 * TA3F0S3 ) / TA3F0"
  requested_calculations["A4F0S"] = "( 0/4 * TA4F0S0 + 1/4 * TA4F0S1 + 2/4 * TA4F0S2 + 3/4 * TA4F0S3 + 4/4 * TA4F0S4 ) / TA4"
  # requested_calculations["A1FS"] = "( 0/1 * TA1FS0 + 1/1 * TA1FS1 ) / TA1F"
  requested_calculations["A2FS"] = "( 0/2 * TA2FS0 + 1/2 * TA2FS1 + 2/2 * TA2FS2 ) / TA2F"
  requested_calculations["A3FS"] = "( 0/3 * TA3FS0 + 1/3 * TA3FS1 + 2/3 * TA3FS2 + 3/3 * TA3FS3 ) / TA3F"
  requested_calculations["A4FS"] = "( 0/4 * TA4FS0 + 1/4 * TA4FS1 + 2/4 * TA4FS2 + 3/4 * TA4FS3 + 4/4 * TA4FS4 ) / TA4"
  
  # Alpha2,3-linked sialylation
  requested_calculations["A1L"] = "( 0/1 * TA1L0 + 1/1 * TA1L1 ) / TA1"
  requested_calculations["A2L"] = "( 0/2 * TA2L0 + 1/2 * TA2L1 + 2/2 * TA2L2 ) / TA2"
  requested_calculations["A3L"] = "( 0/3 * TA3L0 + 1/3 * TA3L1 + 2/3 * TA3L2 + 3/3 * TA3L3 ) / TA3"
  requested_calculations["A4L"] = "( 0/4 * TA4L0 + 1/4 * TA4L1 + 2/4 * TA4L2 + 3/4 * TA4L3 + 4/4 * TA4L4 ) / TA4"
  # requested_calculations["A1Gl"] = "( 0/1 * TA1Gl0 + 1/1 * TA1Gl1 ) / TA1"
  # requested_calculations["A2Gl"] = "( 0/2 * TA2Gl0 + 1/2 * TA2Gl1 + 2/2 * TA2Gl2 ) / TA2"
  # requested_calculations["A3Gl"] = "( 0/3 * TA3Gl0 + 1/3 * TA3Gl1 + 2/3 * TA3Gl2 + 3/3 * TA3Gl3 ) / TA3"
  # requested_calculations["A4Gl"] = "( 0/4 * TA4Gl0 + 1/4 * TA4Gl1 + 2/4 * TA4Gl2 + 3/4 * TA4Gl3 + 4/4 * TA4Gl4 ) / TA4"
  # requested_calculations["A1F0L"] = "( 0/1 * TA1F0L0 + 1/1 * TA1F0L1 ) / TA1F0"
  # requested_calculations["A2F0L"] = "( 0/2 * TA2F0L0 + 1/2 * TA2F0L1 + 2/2 * TA2F0L2 ) / TA2F0"
  # requested_calculations["A3F0L"] = "( 0/3 * TA3F0L0 + 1/3 * TA3F0L1 + 2/3 * TA3F0L2 + 3/3 * TA3F0L3 ) / TA3F0"
  # requested_calculations["A4F0L"] = "( 0/4 * TA4F0L0 + 1/4 * TA4F0L1 + 2/4 * TA4F0L2 + 3/4 * TA4F0L3 + 4/4 * TA4F0L4 ) / TA4"
  # requested_calculations["A1FL"] = "( 0/1 * TA1FL0 + 1/1 * TA1FL1 ) / TA1F"
  # requested_calculations["A2FL"] = "( 0/2 * TA2FL0 + 1/2 * TA2FL1 + 2/2 * TA2FL2 ) / TA2F"
  # requested_calculations["A3FL"] = "( 0/3 * TA3FL0 + 1/3 * TA3FL1 + 2/3 * TA3FL2 + 3/3 * TA3FL3 ) / TA3F"
  # requested_calculations["A4FL"] = "( 0/4 * TA4FL0 + 1/4 * TA4FL1 + 2/4 * TA4FL2 + 3/4 * TA4FL3 + 4/4 * TA4FL4 ) / TA4"
  # requested_calculations["AL"] = "( ( 0/1 * TA1L0 + 1/1 * TA1L1 ) / TA1 + ( 0/2 * TA2L0 + 1/2 * TA2L1 + 2/2 * TA2L2 ) / TA2 + ( 0/3 * TA3L0 + 1/3 * TA3L1 + 2/3 * TA3L2 + 3/3 * TA3L3 ) / TA3 + ( 0/4 * TA4L0 + 1/4 * TA4L1 + 2/4 * TA4L2 + 3/4 * TA4L3 + 4/4 * TA4L4 ) / TA4 ) / ( ( 0/1 * TA1L0 + 1/1 * TA1L1 ) / TA1 + ( 0/2 * TA2L0 + 1/2 * TA2L1 + 2/2 * TA2L2 ) / TA2 + ( 0/3 * TA3L0 + 1/3 * TA3L1 + 2/3 * TA3L2 + 3/3 * TA3L3 ) / TA3 + ( 0/4 * TA4L0 + 1/4 * TA4L1 + 2/4 * TA4L2 + 3/4 * TA4L3 + 4/4 * TA4L4 ) / TA4 + ( 0/1 * TA1E0 + 1/1 * TA1E1 ) / TA1 + ( 0/2 * TA2E0 + 1/2 * TA2E1 + 2/2 * TA2E2 ) / TA2 + ( 0/3 * TA3E0 + 1/3 * TA3E1 + 2/3 * TA3E2 + 3/3 * TA3E3 ) / TA3 + ( 0/4 * TA4E0 + 1/4 * TA4E1 + 2/4 * TA4E2 + 3/4 * TA4E3 + 4/4 * TA4E4 ) / TA4 )"
  #This last trait includes all sialic acids, while the main changes in Gabriane's project seemed to be A3 and A4-linked (multi-antennary glycans)
  
  # Alpha2,6-linked sialylation
  requested_calculations["A1E"] = "( 0/1 * TA1E0 + 1/1 * TA1E1 ) / TA1"
  requested_calculations["A2E"] = "( 0/2 * TA2E0 + 1/2 * TA2E1 + 2/2 * TA2E2 ) / TA2"
  requested_calculations["A3E"] = "( 0/3 * TA3E0 + 1/3 * TA3E1 + 2/3 * TA3E2 + 3/3 * TA3E3 ) / TA3"
  requested_calculations["A4E"] = "( 0/4 * TA4E0 + 1/4 * TA4E1 + 2/4 * TA4E2 + 3/4 * TA4E3 + 4/4 * TA4E4 ) / TA4"
  # requested_calculations["A1Ge"] = "( 0/1 * TA1Ge0 + 1/1 * TA1Ge1 ) / TA1"
  # requested_calculations["A2Ge"] = "( 0/2 * TA2Ge0 + 1/2 * TA2Ge1 + 2/2 * TA2Ge2 ) / TA2"
  # requested_calculations["A3Ge"] = "( 0/3 * TA3Ge0 + 1/3 * TA3Ge1 + 2/3 * TA3Ge2 + 3/3 * TA3Ge3 ) / TA3"
  # requested_calculations["A4Ge"] = "( 0/4 * TA4Ge0 + 1/4 * TA4Ge1 + 2/4 * TA4Ge2 + 3/4 * TA4Ge3 + 4/4 * TA4Ge4 ) / TA4"
  # requested_calculations["A1F0E"] = "( 0/1 * TA1F0E0 + 1/1 * TA1F0E1 ) / TA1F0"
  # requested_calculations["A2F0E"] = "( 0/2 * TA2F0E0 + 1/2 * TA2F0E1 + 2/2 * TA2F0E2 ) / TA2F0"
  # requested_calculations["A3F0E"] = "( 0/3 * TA3F0E0 + 1/3 * TA3F0E1 + 2/3 * TA3F0E2 + 3/3 * TA3F0E3 ) / TA3F0"
  # requested_calculations["A4F0E"] = "( 0/4 * TA4F0E0 + 1/4 * TA4F0E1 + 2/4 * TA4F0E2 + 3/4 * TA4F0E3 + 4/4 * TA4F0E4 ) / TA4"
  # requested_calculations["A1FE"] = "( 0/1 * TA1FE0 + 1/1 * TA1FE1 ) / TA1F"
  # requested_calculations["A2FE"] = "( 0/2 * TA2FE0 + 1/2 * TA2FE1 + 2/2 * TA2FE2 ) / TA2F"
  # requested_calculations["A3FE"] = "( 0/3 * TA3FE0 + 1/3 * TA3FE1 + 2/3 * TA3FE2 + 3/3 * TA3FE3 ) / TA3F"
  # requested_calculations["A4FE"] = "( 0/4 * TA4FE0 + 1/4 * TA4FE1 + 2/4 * TA4FE2 + 3/4 * TA4FE3 + 4/4 * TA4FE4 ) / TA4"
  # requested_calculations["AE"] = "( ( 0/1 * TA1E0 + 1/1 * TA1E1 ) / TA1 + ( 0/2 * TA2E0 + 1/2 * TA2E1 + 2/2 * TA2E2 ) / TA2 + ( 0/3 * TA3E0 + 1/3 * TA3E1 + 2/3 * TA3E2 + 3/3 * TA3E3 ) / TA3 + ( 0/4 * TA4E0 + 1/4 * TA4E1 + 2/4 * TA4E2 + 3/4 * TA4E3 + 4/4 * TA4E4 ) / TA4 ) / ( ( 0/1 * TA1L0 + 1/1 * TA1L1 ) / TA1 + ( 0/2 * TA2L0 + 1/2 * TA2L1 + 2/2 * TA2L2 ) / TA2 + ( 0/3 * TA3L0 + 1/3 * TA3L1 + 2/3 * TA3L2 + 3/3 * TA3L3 ) / TA3 + ( 0/4 * TA4L0 + 1/4 * TA4L1 + 2/4 * TA4L2 + 3/4 * TA4L3 + 4/4 * TA4L4 ) / TA4 + ( 0/1 * TA1E0 + 1/1 * TA1E1 ) / TA1 + ( 0/2 * TA2E0 + 1/2 * TA2E1 + 2/2 * TA2E2 ) / TA2 + ( 0/3 * TA3E0 + 1/3 * TA3E1 + 2/3 * TA3E2 + 3/3 * TA3E3 ) / TA3 + ( 0/4 * TA4E0 + 1/4 * TA4E1 + 2/4 * TA4E2 + 3/4 * TA4E3 + 4/4 * TA4E4 ) / TA4 )"
  #This last trait includes all sialic acids, while the main changes in Gabriane's project seemed to be A3 and A4-linked (multi-antennary glycans)
  
  # Sialylation per galactose
  # requested_calculations["A1GS"] = "( ( 0/1 * TA1S0 + 1/1 * TA1S1 ) / TA1 ) / ( ( 0/1 * TA1G0 + 1/1 * TA1G1 ) / TA1 )"
  # requested_calculations["A2GS"] = "( ( 0/2 * TA2S0 + 1/2 * TA2S1 + 2/2 * TA2S2 ) / TA2 ) / ( ( 0/2 * TA2G0 + 1/2 * TA2G1 + 2/2 * TA2G2 ) / TA2 )"
  # requested_calculations["A3GS"] = "( ( 0/3 * TA3S0 + 1/3 * TA3S1 + 2/3 * TA3S2 + 3/3 * TA3S3 ) / TA3 ) / ( ( 0/3 * TA3G0 + 1/3 * TA3G1 + 2/3 * TA3G2 + 3/3 * TA3G3 ) / TA3 )"
  # requested_calculations["A4GS"] = "( ( 0/4 * TA4S0 + 1/4 * TA4S1 + 2/4 * TA4S2 + 3/4 * TA4S3 + 4/4 * TA4S4 ) / TA4 ) / ( ( 0/4 * TA4G0 + 1/4 * TA4G1 + 2/4 * TA4G2 + 3/4 * TA4G3 + 4/4 * TA4G4 ) / TA4 )"
  # requested_calculations["A1F0GS"] = "( ( 0/1 * TA1F0S0 + 1/1 * TA1F0S1 ) / TA1F0 ) / ( ( 0/1 * TA1F0G0 + 1/1 * TA1F0G1 ) / TA1F0 )"
  # requested_calculations["A2F0GS"] = "( ( 0/2 * TA2F0S0 + 1/2 * TA2F0S1 + 2/2 * TA2F0S2 ) / TA2F0 ) / ( ( 0/2 * TA2F0G0 + 1/2 * TA2F0G1 + 2/2 * TA2F0G2 ) / TA2F0 )"
  # requested_calculations["A3F0GS"] = "( ( 0/3 * TA3F0S0 + 1/3 * TA3F0S1 + 2/3 * TA3F0S2 + 3/3 * TA3F0S3 ) / TA3F0 ) / ( ( 0/3 * TA3F0G0 + 1/3 * TA3F0G1 + 2/3 * TA3F0G2 + 3/3 * TA3F0G3 ) / TA3F0 )"
  # requested_calculations["A4F0GS"] = "( ( 0/4 * TA4F0S0 + 1/4 * TA4F0S1 + 2/4 * TA4F0S2 + 3/4 * TA4F0S3 + 4/4 * TA4F0S4 ) / TA4 ) / ( ( 0/4 * TA4F0G0 + 1/4 * TA4F0G1 + 2/4 * TA4F0G2 + 3/4 * TA4F0G3 + 4/4 * TA4F0G4 ) / TA4 )"
  # requested_calculations["A1FGS"] = "( ( 0/1 * TA1FS0 + 1/1 * TA1FS1 ) / TA1F ) / ( ( 0/1 * TA1FG0 + 1/1 * TA1FG1 ) / TA1F )"
  # requested_calculations["A2FGS"] = "( ( 0/2 * TA2FS0 + 1/2 * TA2FS1 + 2/2 * TA2FS2 ) / TA2F ) / ( ( 0/2 * TA2FG0 + 1/2 * TA2FG1 + 2/2 * TA2FG2 ) / TA2F )"
  # requested_calculations["A3FGS"] = "( ( 0/3 * TA3FS0 + 1/3 * TA3FS1 + 2/3 * TA3FS2 + 3/3 * TA3FS3 ) / TA3F ) / ( ( 0/3 * TA3FG0 + 1/3 * TA3FG1 + 2/3 * TA3FG2 + 3/3 * TA3FG3 ) / TA3F )"
  # requested_calculations["A4FGS"] = "( ( 0/4 * TA4FS0 + 1/4 * TA4FS1 + 2/4 * TA4FS2 + 3/4 * TA4FS3 + 4/4 * TA4FS4 ) / TA4 ) / ( ( 0/4 * TA4FG0 + 1/4 * TA4FG1 + 2/4 * TA4FG2 + 3/4 * TA4FG3 + 4/4 * TA4FG4 ) / TA4 )"
  
  # Alpha2,3-linked sialylation per galactose
  # requested_calculations["A1GL"] = "( ( 0/1 * TA1L0 + 1/1 * TA1L1 ) / TA1 ) / ( ( 0/1 * TA1G0 + 1/1 * TA1G1 ) / TA1 )"
  # requested_calculations["A2GL"] = "( ( 0/2 * TA2L0 + 1/2 * TA2L1 + 2/2 * TA2L2 ) / TA2 ) / ( ( 0/2 * TA2G0 + 1/2 * TA2G1 + 2/2 * TA2G2 ) / TA2 )"
  # requested_calculations["A3GL"] = "( ( 0/3 * TA3L0 + 1/3 * TA3L1 + 2/3 * TA3L2 + 3/3 * TA3L3 ) / TA3 ) / ( ( 0/3 * TA3G0 + 1/3 * TA3G1 + 2/3 * TA3G2 + 3/3 * TA3G3 ) / TA3 )"
  # requested_calculations["A4GL"] = "( ( 0/4 * TA4L0 + 1/4 * TA4L1 + 2/4 * TA4L2 + 3/4 * TA4L3 + 4/4 * TA4L4 ) / TA4 ) / ( ( 0/4 * TA4G0 + 1/4 * TA4G1 + 2/4 * TA4G2 + 3/4 * TA4G3 + 4/4 * TA4G4 ) / TA4 )"
  # requested_calculations["A1F0GL"] = "( ( 0/1 * TA1F0L0 + 1/1 * TA1F0L1 ) / TA1F0 ) / ( ( 0/1 * TA1F0G0 + 1/1 * TA1F0G1 ) / TA1F0 )"
  # requested_calculations["A2F0GL"] = "( ( 0/2 * TA2F0L0 + 1/2 * TA2F0L1 + 2/2 * TA2F0L2 ) / TA2F0 ) / ( ( 0/2 * TA2F0G0 + 1/2 * TA2F0G1 + 2/2 * TA2F0G2 ) / TA2F0 )"
  # requested_calculations["A3F0GL"] = "( ( 0/3 * TA3F0L0 + 1/3 * TA3F0L1 + 2/3 * TA3F0L2 + 3/3 * TA3F0L3 ) / TA3F0 ) / ( ( 0/3 * TA3F0G0 + 1/3 * TA3F0G1 + 2/3 * TA3F0G2 + 3/3 * TA3F0G3 ) / TA3F0 )"
  # requested_calculations["A4F0GL"] = "( ( 0/4 * TA4F0L0 + 1/4 * TA4F0L1 + 2/4 * TA4F0L2 + 3/4 * TA4F0L3 + 4/4 * TA4F0L4 ) / TA4 ) / ( ( 0/4 * TA4F0G0 + 1/4 * TA4F0G1 + 2/4 * TA4F0G2 + 3/4 * TA4F0G3 + 4/4 * TA4F0G4 ) / TA4 )"
  # requested_calculations["A1FGL"] = "( ( 0/1 * TA1FL0 + 1/1 * TA1FL1 ) / TA1F ) / ( ( 0/1 * TA1FG0 + 1/1 * TA1FG1 ) / TA1F )"
  # requested_calculations["A2FGL"] = "( ( 0/2 * TA2FL0 + 1/2 * TA2FL1 + 2/2 * TA2FL2 ) / TA2F ) / ( ( 0/2 * TA2FG0 + 1/2 * TA2FG1 + 2/2 * TA2FG2 ) / TA2F )"
  # requested_calculations["A3FGL"] = "( ( 0/3 * TA3FL0 + 1/3 * TA3FL1 + 2/3 * TA3FL2 + 3/3 * TA3FL3 ) / TA3F ) / ( ( 0/3 * TA3FG0 + 1/3 * TA3FG1 + 2/3 * TA3FG2 + 3/3 * TA3FG3 ) / TA3F )"
  # requested_calculations["A4FGL"] = "( ( 0/4 * TA4FL0 + 1/4 * TA4FL1 + 2/4 * TA4FL2 + 3/4 * TA4FL3 + 4/4 * TA4FL4 ) / TA4 ) / ( ( 0/4 * TA4FG0 + 1/4 * TA4FG1 + 2/4 * TA4FG2 + 3/4 * TA4FG3 + 4/4 * TA4FG4 ) / TA4 )"
  
  # Alpha2,6-linked sialylation per galactose
  # requested_calculations["A1GE"] = "( ( 0/1 * TA1E0 + 1/1 * TA1E1 ) / TA1 ) / ( ( 0/1 * TA1G0 + 1/1 * TA1G1 ) / TA1 )"
  # requested_calculations["A2GE"] = "( ( 0/2 * TA2E0 + 1/2 * TA2E1 + 2/2 * TA2E2 ) / TA2 ) / ( ( 0/2 * TA2G0 + 1/2 * TA2G1 + 2/2 * TA2G2 ) / TA2 )"
  # requested_calculations["A3GE"] = "( ( 0/3 * TA3E0 + 1/3 * TA3E1 + 2/3 * TA3E2 + 3/3 * TA3E3 ) / TA3 ) / ( ( 0/3 * TA3G0 + 1/3 * TA3G1 + 2/3 * TA3G2 + 3/3 * TA3G3 ) / TA3 )"
  # requested_calculations["A4GE"] = "( ( 0/4 * TA4E0 + 1/4 * TA4E1 + 2/4 * TA4E2 + 3/4 * TA4E3 + 4/4 * TA4E4 ) / TA4 ) / ( ( 0/4 * TA4G0 + 1/4 * TA4G1 + 2/4 * TA4G2 + 3/4 * TA4G3 + 4/4 * TA4G4 ) / TA4 )"
  # requested_calculations["A1F0GE"] = "( ( 0/1 * TA1F0E0 + 1/1 * TA1F0E1 ) / TA1F0 ) / ( ( 0/1 * TA1F0G0 + 1/1 * TA1F0G1 ) / TA1F0 )"
  # requested_calculations["A2F0GE"] = "( ( 0/2 * TA2F0E0 + 1/2 * TA2F0E1 + 2/2 * TA2F0E2 ) / TA2F0 ) / ( ( 0/2 * TA2F0G0 + 1/2 * TA2F0G1 + 2/2 * TA2F0G2 ) / TA2F0 )"
  # requested_calculations["A3F0GE"] = "( ( 0/3 * TA3F0E0 + 1/3 * TA3F0E1 + 2/3 * TA3F0E2 + 3/3 * TA3F0E3 ) / TA3F0 ) / ( ( 0/3 * TA3F0G0 + 1/3 * TA3F0G1 + 2/3 * TA3F0G2 + 3/3 * TA3F0G3 ) / TA3F0 )"
  # requested_calculations["A4F0GE"] = "( ( 0/4 * TA4F0E0 + 1/4 * TA4F0E1 + 2/4 * TA4F0E2 + 3/4 * TA4F0E3 + 4/4 * TA4F0E4 ) / TA4 ) / ( ( 0/4 * TA4F0G0 + 1/4 * TA4F0G1 + 2/4 * TA4F0G2 + 3/4 * TA4F0G3 + 4/4 * TA4F0G4 ) / TA4 )"
  # requested_calculations["A1FGE"] = "( ( 0/1 * TA1FE0 + 1/1 * TA1FE1 ) / TA1F ) / ( ( 0/1 * TA1FG0 + 1/1 * TA1FG1 ) / TA1F )"
  # requested_calculations["A2FGE"] = "( ( 0/2 * TA2FE0 + 1/2 * TA2FE1 + 2/2 * TA2FE2 ) / TA2F ) / ( ( 0/2 * TA2FG0 + 1/2 * TA2FG1 + 2/2 * TA2FG2 ) / TA2F )"
  # requested_calculations["A3FGE"] = "( ( 0/3 * TA3FE0 + 1/3 * TA3FE1 + 2/3 * TA3FE2 + 3/3 * TA3FE3 ) / TA3F ) / ( ( 0/3 * TA3FG0 + 1/3 * TA3FG1 + 2/3 * TA3FG2 + 3/3 * TA3FG3 ) / TA3F )"
  # requested_calculations["A4FGE"] = "( ( 0/4 * TA4FE0 + 1/4 * TA4FE1 + 2/4 * TA4FE2 + 3/4 * TA4FE3 + 4/4 * TA4FE4 ) / TA4 ) / ( ( 0/4 * TA4FG0 + 1/4 * TA4FG1 + 2/4 * TA4FG2 + 3/4 * TA4FG3 + 4/4 * TA4FG4 ) / TA4 )"
  
  # New traits added for breast cancer cohort
  # requested_calculations["A34F"] = "( TA3F + TA4F ) / ( TA3 + TA4 )"
  # requested_calculations["CA"] = "( 1 * TA1 + 2 * TA2 + 3 * TA3 + 4 * TA4 ) / TC" # Average number of antennas in the complex type glycans
  # #requested_calculations["THyM"] = "( THy + TM ) / T"
  # requested_calculations["A34L"] = "( 0/3 * TA3L0 + 1/3 * TA3L1 + 2/3 * TA3L2 + 3/3 * TA3L3 + 0/4 * TA4L0 + 1/4 * TA4L1 + 2/4 * TA4L2 + 3/4 * TA4L3 + 4/4 * TA4L4 ) / ( TA3 + TA4 )"
  # requested_calculations["A34E"] = "( 0/3 * TA3E0 + 1/3 * TA3E1 + 2/3 * TA3E2 + 3/3 * TA3E3 + 0/4 * TA4E0 + 1/4 * TA4E1 + 2/4 * TA4E2 + 3/4 * TA4E3 + 4/4 * TA4E4 ) / ( TA3 + TA4 )"
  
  # New traits added for Visceral cohort
  #requested_calculations["A34EL"] = "( TA3E + TA4E ) / ( TA3L + TA4L )"
  #requested_calculations["A34LE"] = "( TA3L + TA4L ) / ( TA3E + TA4E )"
  #requested_calculations["TA34L"] = "(( 0/3 * TA3L0 + 1/3 * TA3L1 + 2/3 * TA3L2 + 3/3 * TA3L3 + 0/4 * TA4L0 + 1/4 * TA4L1 + 2/4 * TA4L2 + 3/4 * TA4L3 + 4/4 * TA4L4 ) / ( TA3 + TA4 )) / ((( 0/3 * TA3L0 + 1/3 * TA3L1 + 2/3 * TA3L2 + 3/3 * TA3L3 + 0/4 * TA4L0 + 1/4 * TA4L1 + 2/4 * TA4L2 + 3/4 * TA4L3 + 4/4 * TA4L4 ) / ( TA3 + TA4 )) + (( 0/3 * TA3E0 + 1/3 * TA3E1 + 2/3 * TA3E2 + 3/3 * TA3E3 + 0/4 * TA4E0 + 1/4 * TA4E1 + 2/4 * TA4E2 + 3/4 * TA4E3 + 4/4 * TA4E4 ) / ( TA3 + TA4 )))"
  #requested_calculations["TA34E"] = "(( 0/3 * TA3E0 + 1/3 * TA3E1 + 2/3 * TA3E2 + 3/3 * TA3E3 + 0/4 * TA4E0 + 1/4 * TA4E1 + 2/4 * TA4E2 + 3/4 * TA4E3 + 4/4 * TA4E4 ) / ( TA3 + TA4 )) / ((( 0/3 * TA3L0 + 1/3 * TA3L1 + 2/3 * TA3L2 + 3/3 * TA3L3 + 0/4 * TA4L0 + 1/4 * TA4L1 + 2/4 * TA4L2 + 3/4 * TA4L3 + 4/4 * TA4L4 ) / ( TA3 + TA4 )) + (( 0/3 * TA3E0 + 1/3 * TA3E1 + 2/3 * TA3E2 + 3/3 * TA3E3 + 0/4 * TA4E0 + 1/4 * TA4E1 + 2/4 * TA4E2 + 3/4 * TA4E3 + 4/4 * TA4E4 ) / ( TA3 + TA4 )))"
  #requested_calculations["TA34S"] = "(( 0/3 * TA3L0 + 1/3 * TA3L1 + 2/3 * TA3L2 + 3/3 * TA3L3 + 0/4 * TA4L0 + 1/4 * TA4L1 + 2/4 * TA4L2 + 3/4 * TA4L3 + 4/4 * TA4L4 ) / ( TA3 + TA4 )) + (( 0/3 * TA3E0 + 1/3 * TA3E1 + 2/3 * TA3E2 + 3/3 * TA3E3 + 0/4 * TA4E0 + 1/4 * TA4E1 + 2/4 * TA4E2 + 3/4 * TA4E3 + 4/4 * TA4E4 ) / ( TA3 + TA4 ))"
  
  # IgG traits
  #requested_calculations["total_IgG_galactosylation"] = "(( H5N4F1 + H5N5F1 + H5N4 + H5N4F1E1 + H5N4L1 + H5N4E1 + H5N4F1E2 + H5N4F1L1E1 ) + ( 1/2 * ( H4N4F1 + H4N5F1 + H4N4 + H4N5 + H4N4F1E1 + H4N5F1E1 + H4N5E1 ))) / T "
  #requested_calculations["IgG1_bisection"] = "( H3N5F1 + H4N5F1 + H5N5F1 + H3N5 + H4N5 + H5N5 + H4N5F1E1 + H5N5F1E1 + H4N5E1 ) / T "
```

## Functions

```{r functions, include=TRUE}
split_nums_chars = function(string){
  # Splits a string into subsequent number and letter blocks
  
  chars = c()
  nums = c()
  last_type = "None"  
  for (i in seq(nchar(string))){
    
    substring = substr(string,i,i)
    
    if (grepl("[0-9]",substring)){
      if (last_type == "number"){
        nums[length(nums)] = paste(nums[length(nums)],substring,sep="")
      }
      else {
        nums = c(nums,substring)
        last_type = "number"
      }  
    }
    else {
      if (last_type == "letter"){
        chars[length(chars)] = paste(chars[length(chars)],substring,sep="")
      }
      else {
        chars = c(chars,substring)
        last_type = "letter"
      }
    } 
  }
  
  output = c()
  output$chars = chars
  output$nums = nums
  
  return(output) 
}

getmatrix_blocks_per_trait = function(traits){
  # Returns matrix of blocks found in traits
  
  blocks = c("F","H","N","S","L","E","Gl","Ge","Ac","P","eGlcA")
  
  matrix_output = matrix(0,length(traits),length(blocks))
  colnames(matrix_output) = blocks
  rownames(matrix_output) = traits  
  
  for(trait in traits){
    trait_split = split_nums_chars(trait)
    for (i in seq(length(trait_split$chars))){
      char = trait_split$chars[i]
      num = as.numeric(trait_split$nums[i])  
      matrix_output[trait,char] = num
    }
  }
  
  return(matrix_output)
}

trait_has_prop = function(matrix_traits,trait,prop){
  # Return 1 if trait matches a property. Else return 0.
  
  p = prop
  t = matrix_traits[trait,]
  result = 0
  
  ### Overall glycan type
  if (p == "T"){
    result = 1
  } 
  else if (p == "PM"){
    if (t["H"] == 3 && t["N"] == 2){
      result = 1
    }
  } 
  else if (p == "U"){
    if ((t["H"] == 3 && t["N"] == 3) ||
        (t["N"] == 3 && t["H"] <= 4 && t["S"] + t["E"] + t["L"] + t["Ge"] + t["Gl"] == 0)){
      result = 1
    }
  }
  else if (p == "M"){
    if (t["H"] > 3 && t["N"] == 2){
      result = 1
    }
  }
  else if (p == "Hy"){
    if(taxonomy=="human"){
      if (t["H"] > t["N"] + 1 && t["N"] > 2 && t["H"] <= t["N"] + 3){
            result = 1
      }
    }
    else {
      if ((t["H"] > t["N"] + 1 && t["N"] == 3 && t["H"] <= 9) || (t["H"] > t["N"] + 1 && t["N"] == 4 && t["H"] > 6 && t["H"] <= 9)){
        result = 1
      }
    }
  }
  else if (p == "C"){
    if(taxonomy=="human"){
      if ((t["N"] > 3 && t["H"] < t["N"] + 2) || (t["N"] == 3 && t["H"] <= 4 && t["S"] + t["E"] + t["L"] + t["Ge"] + t["Gl"] >= 1)){
        result = 1
      }
    } else{
        if ((t["N"] > 3 && t["H"] < ((t["N"] - 2)*2)+4) || (t["N"] == 3 && t["H"] <= 4 && t["S"] + t["E"] + t["L"] + t["Ge"] + t["Gl"] >= 1)){
        result = 1
      }
    }
  }
  
  ### Mannosity
  else if (p == paste("M",gsub(".*?([0-9]+)$", "\\1", p),sep="") && as.numeric(gsub(".*?([0-9]+)$", "\\1", p))>3){
    if (t["N"] == 2 && t["H"] == as.numeric(gsub(".*?([0-9]+)$", "\\1", p))){
      result = 1
    }
  }
  
  ### Antennarity
  else if (p == "B0"){
    if (t["N"] == 4){
      result = 1
    }
  }
  else if (p == "B"){
    if (t["N"] == 5 && t["H"] <= t["N"]){
      result = 1
    }
  }  
  else if (p == "A1"){
    if (t["N"] == 3){
      result = 1
    }
  }  
  else if (p == "A2"){
    if (t["N"] == 4 || (t["N"] == 5 && t["H"] <= t["N"])){
      result = 1
    }
  }   
  else if (p == "A3"){
    if (t["N"] == 5 && !(t["N"] == 5 && t["H"] <= t["N"])){
      result = 1
    }
  }  
  else if (p == "A4"){
    if (t["N"] == 6){
      result = 1
    }    
  }
  else if (p == "ALN"){
    if (t["N"] > 6 && t["H"] == t["N"] + 1){
      result = 1
    }
  }
  
  ### Galactosylation
  else if (p == "G0"){
    if (t["N"] >= 3 && t["H"] == 3 && t["S"] + t["E"] + t["L"] + t["Ge"] + t["Gl"] == 0){
      result = 1
    }    
  }
  else if (p == "G1"){
    if ((t["N"] >= 3 && t["H"] == 4) || (t["N"] == 3 && t["S"] + t["E"] + t["L"] + t["Ge"] + t["Gl"] != 0) ||
          (t["N"] == 3 && t["H"] >= 5)){
      result = 1
    }    
  }
  else if (p == "G2"){
    if ((t["N"] >= 4 && t["H"] == 5) || (t["N"] == 4 && t["S"] + t["E"] + t["L"] + t["Ge"] + t["Gl"] == 2)) {
      result = 1
    }    
  }
  else if (p == "G3"){
    if (t["N"] >= 5 && t["H"] == 6){
      result = 1
    }    
  }
  else if (p == "G4"){
    if (t["N"] >= 6 && t["H"] == 7){
      result = 1
    }   
  }
  
  else if (p == "G"){
    if ((t["H"] >= t["N"] && t["N"] > 2) || (t["S"] + t["E"] + t["L"] + t["Ge"] + t["Gl"] >= 1) ||
          (t["N"] == 5 && t["H"] == 4)){
      result = 1
    }    
  }
  
  ### Sialylation
  # Total sialylation
  # S0-S4
  else if (p == "S0"){
    if (t["S"] + t["E"] + t["L"] + t["Ge"] + t["Gl"] == 0){
      result = 1
    }
  }
  else if (p == paste("S",gsub(".*?([1-9]+)$", "\\1", p),sep="")){
    if ((t["S"] == as.numeric(gsub(".*?([1-9]+)$", "\\1", p))) || (t["E"] + t["L"] + t["Ge"] + t["Gl"] == as.numeric(gsub(".*?([1-9]+)$", "\\1", p)))){
      result = 1
    }
  }  
  else if (p == "S"){
    if ((t["S"] >= 1) || (t["E"] + t["L"] + t["Ge"] + t["Gl"] >= 1)){
      result = 1
    }
  }
  
  # Alpha2,3-linked sialylation
  #L0-L4
  else if (p == paste("L",gsub(".*?([0-9]+)$", "\\1", p),sep="")){
    if (t["L"] == as.numeric(gsub(".*?([0-9]+)$", "\\1", p))){
      result = 1
    }
  }
  else if (p == "L"){
    if (t["L"] >= 1){
      result = 1
    }
  }
  else if (p == paste("Gl",gsub(".*?([0-9]+)$", "\\1", p),sep="")){
    if (t["Gl"] == as.numeric(gsub(".*?([0-9]+)$", "\\1", p))){
      result = 1
    }
  }
  else if (p == "Gl"){
    if (t["Gl"] >= 1){
      result = 1
    }
  }
  
  # Alpha2,6-linked sialylation  
  #E0-E4
  else if (p == paste("E",gsub(".*?([0-9]+)$", "\\1", p),sep="")){
    if (t["E"] == as.numeric(gsub(".*?([0-9]+)$", "\\1", p))){
      result = 1
    }
  }
  else if (p == "E"){
    if (t["E"] >= 1){
      result = 1
    }
  }
  else if (p == paste("Ge",gsub(".*?([0-9]+)$", "\\1", p),sep="")){
    if (t["Ge"] == as.numeric(gsub(".*?([0-9]+)$", "\\1", p))){
      result = 1
    }
  }
  else if (p == "Ge"){
    if (t["Ge"] >= 1){
      result = 1
    }
  }
  
  ### Fucosylation
  #"F0"-"F7"
  else if (p == paste("F",gsub(".*?([0-9]+)$", "\\1", p),sep="")){
    if (t["F"] == as.numeric(gsub(".*?([0-9]+)$", "\\1", p))){
      result = 1
    }
  }
  else if (p == "F"){
    if (t["F"] >= 1){
      result = 1
    }
  }
  else if (p == "Fa"){
    if (t["F"] >= 2){
      result = 1
    }
  }
  
  ### Acetylation
  #"Ac0"-"Ac4"
  else if (p == paste("Ac",gsub(".*?([0-9]+)$", "\\1", p),sep="")){
    if (t["Ac"] == as.numeric(gsub(".*?([0-9]+)$", "\\1", p))){
      result = 1
    }
  }
  else if (p == "Ac"){
    if (t["Ac"] >= 1){
      result = 1
    }
  }
  
  return(result)
}

getmatrix_trait_properties = function(matrix_traits){
  # Returns matrix of properties true for traits
  
  props = c("T","PM","U","M","Hy","C",
            "M4","M5","M6","M7","M8","M9","M10","M11","M12",
            "A1","A2","A3","A4","ALN",
            "B0","B",
            "G0","G1","G2","G3","G4","G",
            "S0","S1","S2","S3","S4","S",
            "L0","L1","L2","L3","L4","L",
            "E0","E1","E2","E3","E4","E",
            "Gl0","Gl1","Gl2","Gl3","Gl4","Gl",
            "Ge0","Ge1","Ge2","Ge3","Ge4","Ge",
            "F0","F1","F2","F3","F4","F5","F6","F7","F","Fa",
            "Ac0","Ac1","Ac2","Ac3","Ac4","Ac",
            "eGlcA")
  
  traits = rownames(matrix_traits)
  matrix_output = matrix(0,length(traits),length(props))
  colnames(matrix_output) = props
  rownames(matrix_output) = traits
  
  for (trait in traits){
    for (prop in props){
      matrix_output[trait,prop] = trait_has_prop(matrix_traits,trait,prop)
    }
  }
  return(matrix_output)
}

getmatrix_group_calculations = function(matrix_props){
  # Generate group calculation matrix on basis of properties matrix
  
  m = matrix_props
  names = rownames(matrix_props)
  
  matrix_output = matrix("",0,1)
  
  ### Base
  # T
  sumlist = c()
  for (name in names){
    if (m[name,"T"] == 1){
      sumlist = c(sumlist, name)
    }    
  }
  matrix_output["T"] = paste(sumlist,collapse=" ")
  
  # TP
  sumlist = c()
  for (name in names){
    if (m[name,"PM"] == 1){
      sumlist = c(sumlist, name)
    }    
  }
  matrix_output["TP"] = paste(sumlist,collapse=" ")
  
  # TU
  sumlist = c()
  for (name in names){
    if (m[name,"U"] == 1){
      sumlist = c(sumlist, name)
    }    
  }
  matrix_output["TU"] = paste(sumlist,collapse=" ")
  
  # TM
  sumlist = c()
  for (name in names){
    if (m[name,"M"] == 1){
      sumlist = c(sumlist, name)
    }    
  }
  matrix_output["TM"] = paste(sumlist,collapse=" ")
  
  # THy
  sumlist = c()
  for (name in names){
    if (m[name,"Hy"] == 1){
      sumlist = c(sumlist, name)
    }    
  }
  matrix_output["THy"] = paste(sumlist,collapse=" ")
  
  # TC
  sumlist = c()
  for (name in names){
    if (m[name,"C"] == 1){
      sumlist = c(sumlist, name)
    }    
  }
  matrix_output["TC"] = paste(sumlist,collapse=" ")
  
  # A1-A4
  for (count in seq(1,4)){
    sumlist = c()
    for (name in names){
      if (m[name,paste("A",count,sep="")] == 1){
        sumlist = c(sumlist, name)
      }    
    }
    matrix_output[paste("A",count,sep="")] = paste(sumlist,collapse=" ")
  }
  
  # ALN (LacNac)
  sumlist = c()
  for (name in names){
    if (m[name,"ALN"] == 1){
      sumlist = c(sumlist, name)
    }    
  }
  matrix_output["ALN"] = paste(sumlist,collapse=" ")
  
  # F0-F7
  for (count in seq(0,7)){
    sumlist = c()
    for (name in names){
      if (m[name,paste("F",count,sep="")] == 1){
        sumlist = c(sumlist, name)
      }    
    }
    matrix_output[paste("F",count,sep="")] = paste(sumlist,collapse=" ")
  }
  
  # S0-S4
  for (count in seq(0,4)){
    sumlist = c()
    for (name in names){
      if (m[name,paste("S",count,sep="")] == 1){
        sumlist = c(sumlist, name)
      }    
    }
    matrix_output[paste("S",count,sep="")] = paste(sumlist,collapse=" ")
  }
  
  ### High mannose compositions
  for (count in seq(4,12)){
    sumlist = c()
    for (name in names){
      if (m[name,paste("M",count,sep="")] == 1){
        sumlist = c(sumlist,name)
      }
    }
    matrix_output[paste("TM",count,sep="")] = paste(sumlist,collapse=" ")
  }
  
  # More overall traits
  traits = c("G","S","L","E","Gl","Ge","F","Ac")
  nums = c("0","1","2","3","4","")
  
  for (trait in traits){
    for (num in nums){
      trait_name = paste(trait,num,sep="")
      output_name = paste("T",trait_name,sep="")
      sumlist = c()
      for (name in names){
        if (m[name,trait_name] == 1 && m[name,"C"] == 1){
          sumlist = c(sumlist,name)
        }
      }
      matrix_output[output_name] = paste(sumlist,collapse=" ")      
    }
  } 
  
  # TFa
  sumlist = c()
  for (name in names){
    if (m[name,"Fa"] == 1 && m[name,"C"] == 1){
      sumlist = c(sumlist, name)
    }    
  }
  matrix_output["TFa"] = paste(sumlist,collapse=" ")
  
  ### Bisection
  bisection_list = c("B0","B")
  antennary_list = c("","A2")
  fucosylation_list = c("","F0","F")
  galactosylation_list = c("","G0","G1","G2","G")
  sialylation_list = c("","S0","S1","S2","S")
  for (bisection in bisection_list){
    for (antenna in antennary_list){
      for (fucose in fucosylation_list){
        for (galactose in galactosylation_list){
          for (sialylation in sialylation_list){
            trait_name = paste(antenna,fucose,galactose,sialylation,bisection,sep="")
            output_name = paste("T",trait_name,sep="")
            sumlist = c()
            for (name in names){
              if (m[name,"C"] == 1 && m[name,bisection] == 1 && (antenna == "" || m[name,antenna] == 1) &&
                    (fucose == "" || m[name,fucose] == 1) && (galactose == "" || m[name,galactose] == 1) &&
                    (sialylation == "" || m[name,sialylation] == 1)){
                sumlist = c(sumlist,name)
              }        
            }
            matrix_output[output_name] = paste(sumlist,collapse=" ")
          }
        }
      }
    }
  }
  
  ### The works: all combinations of antennary, fucosylation, galactosylation and sialylation
  antennary_list = c("A1","A2","A3","A4","ALN")
  fucosylation_list = c("","F0","F","Fa")
  galactosylation_list = c("","G0","G1","G2","G3","G4","G")
  sialylation_list = c("","S0","S1","S2","S3","S4","S",
                       "L0","L1","L2","L3","L4","L",
                       "E0","E1","E2","E3","E4","E",
                       "Gl0","Gl1","Gl2","Gl3","Gl4","Gl",
                       "Ge0","Ge1","Ge2","Ge3","Ge4","Ge")
  for (antenna in antennary_list){
    for (fucose in fucosylation_list){
      for (galactose in galactosylation_list){
        for (sialylation in sialylation_list){
          trait_name = paste(antenna,fucose,galactose,sialylation,sep="")
          output_name = paste("T",trait_name,sep="")
          sumlist = c()
          for (name in names){
            if (m[name,"C"] == 1 &&
                  m[name,antenna] == 1 &&
                  (fucose == "" || m[name,fucose] == 1) &&
                  (galactose == "" || m[name,galactose] == 1) &&
                  (sialylation == "" || m[name,sialylation] == 1)){
              sumlist = c(sumlist,name)
            }
          }
          matrix_output[output_name] = paste(sumlist,collapse=" ")
        }
      }
    }
  }
  
  return(matrix_output)
}

get_sumstring = function(string){
  # Create parenthesized sumstring from space delimited string.
  # Returns "(0)" on empty string.
  if (string == ""){
    newstring = "(0)"
  }
  else {
    newstring = "("
    for (substring in unlist(strsplit(string,split=" "))){
      substring = paste(substring,"+")
      newstring = paste(newstring,substring)
    }
    newstring = substr(newstring,1,nchar(newstring)-2)
    newstring = paste(newstring,")")
  }
  
  return(newstring)
}

get_mathstring = function(name, string, matrix_groups){
  # Generate interpretable math string from easy manual string.
  
  newstring = paste(name,"=")
  
  substring_list = unlist(strsplit(string,split=" "))
  for (i in seq(length(substring_list))){
    if (substring_list[i] %in% names(matrix_groups)){
      subsubstring = matrix_groups[substring_list[i]]
      substring_list[i] = get_sumstring(subsubstring)
    }
  }
  
  substring = paste(substring_list,collapse=" ")
  newstring = paste(newstring,substring)
  
  return(newstring)
}

getmatrix_derived_calculations = function(requested_calculations, matrix_groups){
  # Make array with derived trait calculations based on group matrix
  # Whitespace is used for splitting later! Put one around all expressions!
  
  for (name in names(requested_calculations)){
    requested_calculations[name] = get_mathstring(name,requested_calculations[name],matrix_groups)
  }
  
  return(requested_calculations)
}

get_calcstring = function(string,data,glycan_names){
  # Format calculation string into something parsable by eval().
  
  newstring = ""
  data_name = deparse(substitute(data))
  substring_list = unlist(strsplit(string,split=" "))
  
  
  newstring = paste(data_name,".derived[,'",substring_list[1],"']",sep="")
  
  for (substring in substring_list[2:length(substring_list)]){
    if (substring %in% glycan_names){
      substring = paste(data_name,"[,'",substring,"']",sep="")
    }
    newstring = paste(newstring,substring)
  }
  
  return(newstring)  
}
```

## Read data

```{r data import, include=TRUE}
if(str_sub(import_file_name, start= -4) == ".txt"){
  data = read.table(fname_data,sep="\t",header=1,na.string=c(NA,"NA",""),fill=TRUE)
}
if(str_sub(import_file_name, start= -5) == ".xlsx"){
  if(!is.null(sheetName)){
    data <- openxlsx::read.xlsx(fname_data, sheet=sheetName, colNames= TRUE, sep=" ")
  } else {
    data <- openxlsx::read.xlsx(fname_data, colNames= TRUE, sep=" ")
  }
}

start_col <- which(str_detect(colnames(data), "^[Hh]([0-9]+)[Nn]([0-9]+).*$") == TRUE)[1] # column number where the glycans start, starts with HxNy
```

## Calculations

```{r calculations execution, include=TRUE}
# Select glycan species to use for derived traits
glycan_names = names(data)[start_col:ncol(data)]

# Run derived trait background functions
matrix_glycans = getmatrix_blocks_per_trait(glycan_names)
matrix_properties = getmatrix_trait_properties(matrix_glycans)
matrix_groups = getmatrix_group_calculations(matrix_properties)
matrix_derivedmath = as.matrix(getmatrix_derived_calculations(requested_calculations, matrix_groups))
colnames(matrix_derivedmath) <- c("Calculation")
matrix_groups = as.matrix(matrix_groups)
colnames(matrix_groups) <- c("Composition")

data.derived <- data[1:start_col-1]

i=0
# Perform derived trait calculation
for (math in matrix_derivedmath){
  i=i+1
  calc_string = get_calcstring(math,data,glycan_names)
  tryCatch({
    eval(parse(text=calc_string))
  }, error = function(e) {
    calc_string = paste0("data.derived[,'", row.names(matrix_derivedmath)[i], "'] = (0) / data.derived[,'T']")
    eval(parse(text=calc_string))
  })
}
i=0

data.derived[is.na(data.derived)] = NA # this line prevents NA values to be presented as #NUM!
data.derived["T"] = signif(data.derived["T"], digits = 7) # because of rounding T might end up as 0.99999999 instead of 1. This line prevents that by rounding to 7 digits.
data.derived.filter <- data.frame(Sums = colSums(data.derived[,(start_col+1):ncol(data.derived)])!=0 & colSums(data.derived[,(start_col+1):ncol(data.derived)])!=nrow(data.derived) & !is.na(colSums(data.derived[,(start_col+1):ncol(data.derived)])))
data.derived.filter <- rownames(data.derived.filter %>% filter(Sums == TRUE))
data.derived.curated <- cbind(data.derived[, 1:start_col],data.derived[,data.derived.filter])
allData.curated <- cbind(data,data.derived.curated[,data.derived.filter])
```

## Write to file

```{r write to file, include=TRUE}
# Write new data to file
if(str_detect(export_file_name, pattern = fixed(".txt"))){
  # Write auxilary data to files
  write.table(matrix_properties,"composition_properties.txt",sep="\t")
  write.table(matrix_groups,"composition_groups.txt",sep="\t")
  write.table(matrix_derivedmath,"derived_math.txt",sep="\t")
  write.table(data.derived, fname_export,sep="\t",row.names=FALSE)
  write.table(data.derived.curated, paste(fname_export, "_curated", sep=""),sep="\t",row.names=FALSE)
}

if(str_detect(export_file_name, pattern = fixed(".xlsx"))){
  wb <- openxlsx::createWorkbook()
  openxlsx::addWorksheet(wb, "original")
  openxlsx::addWorksheet(wb, "derived_traits")
  openxlsx::addWorksheet(wb, "derived_traits_curated")
  openxlsx::addWorksheet(wb, "all_traits_curated")
  openxlsx::addWorksheet(wb, "composition_properties")
  openxlsx::addWorksheet(wb, "composition_groups")
  openxlsx::addWorksheet(wb, "derived_math")
  openxlsx::writeData(wb, "original", data, rowNames = FALSE, keepNA = TRUE)
  openxlsx::writeData(wb, "derived_traits", data.derived, rowNames = FALSE, keepNA = TRUE)
  openxlsx::writeData(wb, "derived_traits_curated", data.derived.curated, rowNames = FALSE, keepNA = TRUE)
  openxlsx::writeData(wb, "all_traits_curated", allData.curated, rowNames = FALSE, keepNA = TRUE)
  openxlsx::writeData(wb, "composition_properties", matrix_properties, rowNames = TRUE)
  openxlsx::writeData(wb, "composition_groups", matrix_groups, rowNames = TRUE)
  openxlsx::writeData(wb, "derived_math", matrix_derivedmath, rowNames = TRUE)
  openxlsx::saveWorkbook(wb, file = fname_export, overwrite = TRUE)
}
```
Data written to `r fname_export`