---
title: "Glycan Raw data Quality Control"
authors: "Viktoria Dotz, Bas Jansen (v1.0), Elham Memarian (v1.1), Marco Bladergroen (v2.0 and higher)"
date: "version: 4.0.1, 2025-03-06, executed `r  Sys.Date()`"
version: 4.0.1
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
editor_options: 
 chunk_output_type: inline
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding,
 output_file = file.path(dirname(inputFile), paste0('Glycan Data Curation ', format(Sys.time(), '%Y-%m-%d %H_%M_%S'),'.html'))) })
---
# Version history
1.0  : 2019-03-18: Viktoria Dotz, Bas Jansen\
1.1  : 2021-04-13: Elham Memarian\
2.0  : 2021-05-12: Marco Bladergroen: updated from .txt file usage to .xlsx file usage. Introduction of settings block and metaData file.\
2.1  : 2022-05-19: Marco Bladergroen: dots removed from column headers.\
2.2  : 2022-08-05: Marco Bladergroen: error in reference to table for sigmaTotal and sigmaFA corrected.\
2.2.1: 2022-08-16: Marco Bladergroen: Minor layout changes.\
2.2.2: 2022-12-14: Marco Bladergroen: Included option to leave out specific glycans (e.g. DP7 or specific other glycans used for quality control)\
2.3.0: 2023-02-17: Marco Bladergroen: Eliminated the need for SampleTypes. Only SampleType for blanks is kept, which can be either numbered or named; selection and filtering simplified.\
2.4.0: 2023-05-04: Marco Bladergroen: Mass list added to output.\
3.0.0: 2022-10-13: Marco Bladergroen: Analyte curation based on rSD (30%) included.\
3.1.0: 2023-12-19: Marco Bladergroen: Versions 2.x.y and 3.0.0 combined. Error in column shift corrected.\
3.1.1: 2024-01-25: Marco Bladergroen: Small error with 'exclude' corrected.\
3.1.2: 2024-03-15: Marco Bladergroen: Excluded traits (for instance internal standards) included in analyte verdict for quality checks, but excluded later.\
3.1.3: 2024-04-04: Marco Bladergroen: Some typo's corrected after change in Data conversion script (version 11).\
3.1.4: 2025-03-03: Marco Bladergroen: Option to include all samples, without first filtering on sd or IQR (set both to NULL).\
4.0.0: 2025-03-06: Marco Bladergroen: Glycan renormalization now based on passed analytes per cohort.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(tidyverse)
library(openxlsx)
library(ggplot2)
library(knitr)
library(tibble)
```

```{r settings}
wd <- getwd()
dataFileName <- "<Name of datafile after 'Check blanks' script>.xlsx"
sheetNameData <- "summary_wo_blanks"
data_conCol <- "xy_filename" # Type here the name of the column in the datafile connecting the data with the metadata. Also define when no metafile is used = unique identifier for sample
dataFileName <- paste(wd, dataFileName, sep="/")

metaFileName <- "<Name of your metaData file>.xlsx"
sheetNameMeta <- "metaData"
meta_conCol <- "xy_filename" # Type here the name of the column in the metafile connecting the data with the metadata
if(metaFileName!=""){
  metaFileName <- paste(wd, metaFileName, sep="/")
}

outputFileName <- paste(wd, paste0("<Name of the output file>", format(Sys.time(), '%Y-%m-%d %H_%M_%S'), ".xlsx"), sep="/")

# Sample type for blanks (used for filtering)
typeNumBlank <- c(0, "blank", "Blank", "PBS")
typeNumTech <- c(10, "Visucon", "visucon") #Set all possible notations for technical samples (e.g. Visucon) to be included for data checks, but excluded from analysis

colNameCohort <- "Subgroup" # If NULL, all samples will be analyzed as a single cohort (Visucon excluded), otherwise per cohort group.
colNameSampleType <- "SampleType"

# include or exclude specific samples: put names used in data_conCol or meta_conCol here. Use a data frame: first column (vector) are excluded samples, second column (vector) are exclusion reasons; exclude <- data.frame(c(excluded), c(reasons_excluded)). NULL if not needed
include <- NULL #data.frame(c(), c())
exclude <- NULL #data.frame(c(), c())

# list naming
if(!is.null(include)){
  names(include) <- c(data_conCol, "Inclusion reason")
}

if(!is.null(exclude)){
  names(exclude) <- c(data_conCol, "Exclusion reason")
}

# Exclude specific columns (normally used for quality control like DP7 (H7))
excludeTrait <- c("H7")

## Spectra curation cut-offs
sd_cutoff <- 3 #NULL if IQR should be used
IQR_factor <- NULL # select (1.5 or) 3: NULL if mean and xSD should be used

## Analyte curation cutoffs
SN <- 9
IPQ <- 0.2
PPM <- 10
RSD <- NULL #30 # Set to NULL if analyte curation based on rSD not needed. Default=30.
k_Cutoff_per_group <- 0.25 # Fraction of samples that should contain a certain analyte
k_Cutoff_anal_num <- 0.25 # Fraction of number of curated analytes a sample should contain (per biological group) in order to be included
k_Cutoff_TotInt_cur <- 0.01 # Fraction of curated Total Intensity a sample should have (per biological group) in order to be included.

default_color <- "green"
colorSetLUMC <- c("#007cc2", "#c0965c", "#e3004f", "#003c7d", "#009fbd", "#6e90ab", "#80a51b","#f39200", "#cf7a92", "#8d716f", "#7f5ca3", "#ffcc00")
colorSetGerda <- c("#36465f", "#00857c", "#97c8ab", "#bf5349", "#e15a54", "#e97264", "#be9851", "#e5aa2a") # Tumblr (blue), Etos (green), The Correspondent (lightgreen), Wehkamp (red), The Correspondent (mediumred), Google (lightred), Tenslife (brownyellow), Douwe Egberts (yellow), NA (grey)
colorNA <- "#D3D3D3"
figCols <- colorSetLUMC #c("red", "blue", "purple", "green" )
```

### Dataset Loading
```{r Dataset loading}
if(sheetNameData != ""){
  data <- read.xlsx(dataFileName, sheet = sheetNameData, colNames= TRUE, sep=" ")
} else {
  data <- read.xlsx(dataFileName, colNames= TRUE, sep=" ")
}

##Variable subsets
if(metaFileName != ""){
  if(sheetNameMeta != ""){
    metaData <- read.xlsx(metaFileName, sheet = sheetNameMeta, colNames= TRUE, sep=" ")
  } else {
    metaData <- read.xlsx(metaFileName, colNames= TRUE, sep=" ")
  }
  metaData <- metaData %>% mutate(across(where(is.numeric), as.factor))
} else {
  # assume metaData is in data
  metaData <- data[1:(which(colnames(data)=="Total Intensity")-1)]
}

if(ncol(metaData)>1 & length(which(colnames(data) %in% colnames(metaData)))>1){
  allData <- metaData %>% 
        inner_join(y = data[-(which(colnames(data) %in% colnames(metaData))[!colnames(metaData) %in% c(data_conCol)])], by = c(eval(parse(text = paste0("\"", meta_conCol, "\" = \"", data_conCol, "\"")))))
} else if(ncol(metaData)>1){
  allData <- metaData %>% 
        inner_join(y = data, by = c(eval(parse(text = paste0("\"", meta_conCol, "\" = \"", data_conCol, "\"")))))
} else {
  allData <- data
}

names(figCols) <- levels(allData[[colNameSampleType]])
```

### Spectrum Quality Plots before curation
```{r Initial Spectrum Quality Plots}
# Total Intensity vs.Fraction of analyte >9 S/N
QCplot <- ggplot(allData, aes(x=allData[,"Total Intensity"], y=allData[,"Fraction of Analyte above S/N cut-off (9)"], color=!!sym(colNameSampleType))) +
            geom_point(shape=1) +
            scale_colour_manual(limits = levels(unique(allData[[colNameSampleType]])), values=figCols) +
            xlab("Total intensity") + ylab("Fraction of analytes with S/N > 9")
QCplot+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                 panel.background = element_blank(), axis.line = element_line(colour = "black"))

# Total Intensity vs.Fraction of analyte area minus background >9 S/N
QCplot <- ggplot(allData, aes(x=allData[,"Total Intensity"], y=allData[,"Fraction of Analyte Area - Background Area above S/N cut-off (9)"], color=!!sym(colNameSampleType))) +
            geom_point(shape=1) +
            scale_colour_manual(limits = levels(unique(allData[[colNameSampleType]])), values=figCols) +
            xlab("Total intensity") + ylab("Fraction of analyte area minus background with S/N > 9")
QCplot+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                 panel.background = element_blank(), axis.line = element_line(colour = "black"))

# Total Intensity vs.Fraction of spectrum in analytes
QCplot <- ggplot(allData, aes(x=allData[,"Total Intensity"], y=allData[,"Fraction of spectrum in analytes"], color=!!sym(colNameSampleType))) +
            geom_point(shape=1) +
            scale_colour_manual(limits = levels(unique(allData[[colNameSampleType]])), values=figCols) +
            xlab("Total intensity") + ylab("Fraction of spectrum in analytes")
QCplot+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                 panel.background = element_blank(), axis.line = element_line(colour = "black"))
```

# Spectrum-QC based outlier removal (lower end only; after blank removal)
```{r Spectrum QC after blank removal}
inputData <- allData
data_Blnk <- allData %>% filter(!!sym(colNameSampleType) %in% typeNumBlank) # should already been eliminated by R_Readable_data_analysis script. If not empty, blanks are not blank
if(nrow(data_Blnk)>0){
  print("Blanks with signal")
  kable(as.data.frame(data_Blnk))
}
allData <- allData %>% filter(!(!!sym(colNameSampleType) %in% typeNumBlank)) # for blank removal

if(!is.null(sd_cutoff)){
  ## A) Mean and xSD for outlier removal
  # Select QC parameter(s) of interest
  muTotal <- mean(allData$`Total Intensity`,na.rm=TRUE)
  sigmaTotal <- sd(allData$`Total Intensity`,na.rm=TRUE)
  
  muFA <- mean(allData$`Fraction of Analyte above S/N cut-off (9)`,na.rm=TRUE)
  sigmaFA <- sd(allData$`Fraction of Analyte above S/N cut-off (9)`,na.rm=TRUE)

  muFABG <- mean(allData$`Fraction of Analyte Area - Background Area above S/N cut-off (9)`,na.rm=TRUE)
  sigmaFABG <- sd(allData$`Fraction of Analyte Area - Background Area above S/N cut-off (9)`,na.rm=TRUE)

  lowerTotal <- muTotal - sd_cutoff * sigmaTotal
  lowerTotal
  #upperTotal <- muTotal + sd_cutoff * sigmaTotal #maybe don't use?
  
  lowerFA <- muFA - sd_cutoff * sigmaFA
  #upperFA <- muFA + sd_cutoff * sigmaFA #maybe don't use?
  
  lowerFABG <- muFABG - sd_cutoff * sigmaFABG
  #upperFABG <- muFABG + sd_cutoff * sigmaFABG #maybe don't use?

  #Total Intensity, Fraction of Analytes above S/N cut-off(9) and Fraction of Analytes - Background Area above S/N cut-off(9) all above 'lower' values
  data_curated <- allData %>% filter(`Total Intensity`>lowerTotal & `Fraction of Analyte above S/N cut-off (9)`>lowerFA & `Fraction of Analyte Area - Background Area above S/N cut-off (9)`>lowerFABG)
} else if(!is.null(IQR_factor)){
  ## B) IQR rule for outlier removal
  upperTotal <- quantile(allData$`Total Intensity`,0.75, na.rm=TRUE) + (IQR(allData$`Total Intensity`, na.rm=TRUE) * IQR_factor ) #maybe don't use?
  lowerTotal <- quantile(allData$`Total Intensity`,0.25, na.rm=TRUE) - (IQR(allData$`Total Intensity`, na.rm=TRUE) * IQR_factor )
  upperFA <- quantile(allData$`Fraction of Analyte above S/N cut-off (9)`,0.75, na.rm=TRUE) + (IQR(allData$`Fraction of Analyte above S/N cut-off (9)`, na.rm=TRUE) * IQR_factor ) #maybe don't use?
  lowerFA <- quantile(allData$`Fraction of Analyte above S/N cut-off (9)`,0.25, na.rm=TRUE) - (IQR(allData$`Fraction of Analyte above S/N cut-off (9)`, na.rm=TRUE) * IQR_factor )
  upperFABG <- quantile(allData$`Fraction of Analyte Area - Background Area above S/N cut-off (9)`,0.75, na.rm=TRUE) + (IQR(allData$`Fraction of Analyte Area - Background Area above S/N cut-off (9)`, na.rm=TRUE) * IQR_factor ) #maybe don't use?
  lowerFABG <- quantile(allData$`Fraction of Analyte Area - Background Area above S/N cut-off (9)`,0.25, na.rm=TRUE) - (IQR(allData$`Fraction of Analyte Area - Background Area above S/N cut-off (9)`, na.rm=TRUE) * IQR_factor )
 
  #Fraction of Analytes-based outlier removal
  data_curated <- allData %>% filter(`Total Intensity`>lowerTotal && `Total Intensity`<upperTotal && `Fraction of Analyte above S/N cut-off (9)`>lowerFA && `Fraction of Analyte above S/N cut-off (9)`<upperFA && `Fraction of Analyte Area - Background Area above S/N cut-off (9)`>lowerFABG && `Fraction of Analyte Area - Background Area above S/N cut-off (9)`<upperFABG)
} else {
  data_curated <- allData
}
txt <- paste("row.names(data_curated) <- data_curated$", meta_conCol, sep="")
eval(parse(text=txt))

# % of spectra kept
print(paste(round(nrow(data_curated)*100/nrow(allData), 1), "% (", nrow(data_curated), ") of spectra kept, ", nrow(allData)-nrow(data_curated), " removed", sep=""))

# #select only spectra that were kept / removed
txt <- paste("!", meta_conCol, " %in% data_curated$", meta_conCol, sep="")
data_removed <- allData %>% filter(eval(parse(text=txt)))
txt <- paste("row.names(data_removed) <- data_removed$", meta_conCol, sep="")
eval(parse(text=txt))

options(knitr.kable.NA = '')
kable(as.data.frame(head(data_curated) %>% add_row()), caption = "Curated data.")
kable(as.data.frame(data_removed), caption = "Removed data.")

if(!is.null(include)){
  data_curated <- rbind(data_curated, data_removed %>% filter(eval(parse(text=meta_conCol)) %in% include[[data_conCol]]))
  data_removed <- data_removed %>% filter(!eval(parse(text=meta_conCol)) %in% include[[data_conCol]])
}
if(!is.null(exclude)){
  data_removed_manual <- data_curated %>% filter(eval(parse(text=meta_conCol)) %in% exclude[[data_conCol]])
  data_curated <- data_curated %>% filter(!eval(parse(text=meta_conCol)) %in% exclude[[data_conCol]])
}

data_curated <- data_curated %>% droplevels()
data_removed <- data_removed %>% droplevels()

if(!is.null(exclude)){
  data_removed_manual <- data_removed_manual %>% droplevels()
  data_removed_manual <- right_join(exclude, data_removed_manual, data_conCol)
}

wb <- createWorkbook()
addWorksheet(wb, "input_data")
writeData(wb, "input_data", inputData, rowNames = FALSE, keepNA = TRUE)
if(nrow(data_Blnk)>0){
  addWorksheet(wb, "removed_blanks")
  writeData(wb, "removed_blanks", data_Blnk, rowNames = FALSE, keepNA = TRUE)
}
addWorksheet(wb, "curated_spectra_1")
addWorksheet(wb, "removed_spectra_1")
writeData(wb, "curated_spectra_1", data_curated, rowNames = FALSE, keepNA = TRUE)
writeData(wb, "removed_spectra_1", data_removed, rowNames = FALSE, keepNA = TRUE)
if(!is.null(exclude)){
  addWorksheet(wb, "removed_spectra_manual")
  writeData(wb, "removed_spectra_manual", data_removed_manual, rowNames = FALSE, keepNA = TRUE)
}
saveWorkbook(wb, file = outputFileName, overwrite = TRUE)
```

### Curated Spectra quality plots incl. boxplot for Fraction of analyte >9 S/N distribution
```{r Curated Spectra quality plots}
# Total Intensity vs.Fraction of analyte >9 S/N
QCplot <- ggplot(data_curated, aes(x=data_curated[,"Total Intensity"], y=data_curated[,"Fraction of Analyte above S/N cut-off (9)"], color=!!sym(colNameSampleType))) +
            geom_point(shape=1) +
            scale_colour_manual(limits = levels(unique(data_curated[[colNameSampleType]])), values=figCols) +
            xlab("Total intensity") + ylab("Fraction of analytes with S/N > 9")
QCplot+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                 panel.background = element_blank(), axis.line = element_line(colour = "black"))

# Total Intensity vs.Fraction of analyte area minus background >9 S/N
QCplot <- ggplot(data_curated, aes(x=data_curated[,"Total Intensity"], y=data_curated[,"Fraction of Analyte Area - Background Area above S/N cut-off (9)"], color=!!sym(colNameSampleType))) +
            geom_point(shape=1) +
            scale_colour_manual(limits = levels(unique(data_curated[[colNameSampleType]])), values=figCols) +
            xlab("Total intensity") + ylab("Fraction of analyte area minus background with S/N > 9")
QCplot+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                 panel.background = element_blank(), axis.line = element_line(colour = "black"))

# Total Intensity vs.Fraction of spectrum in analytes
QCplot <- ggplot(data_curated, aes(x=data_curated[,"Total Intensity"], y=data_curated[,"Fraction of spectrum in analytes"], color=!!sym(colNameSampleType))) +
            geom_point(shape=1) +
            scale_colour_manual(limits = levels(unique(data_curated[[colNameSampleType]])), values=figCols) +
            xlab("Total intensity") + ylab("Fraction of spectrum in analytes")
QCplot+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                 panel.background = element_blank(), axis.line = element_line(colour = "black"))

#Boxplot
boxplot(data_curated[,"Fraction of Analyte above S/N cut-off (9)"], xlab="spectra kept", ylab="Fraction of Analyte > S/N 9",main="Curated Spectra quality plots")
```

### Analyte quality control
```{r Analyte quality control}
SN_cols <- which(str_detect(colnames(data_curated), "^S/N_") == TRUE) #indicate column numbers where S/N value is stored
IPQ_cols <- which(str_detect(colnames(data_curated), "^IPQ_") == TRUE) #indicate column numbers where IPQ value is stored
AbsPPM_cols <- which(str_detect(colnames(data_curated), "^AbsPPM_") == TRUE) #indicate column numbers where AbsPPM value is stored
analytes_num <- length(which(str_detect(colnames(data_curated), "^S/N_") == TRUE)) #indicate the number of analytes extracted by MassyTools

SN_columns <- colnames(data_curated[SN_cols])
IPQ_columns <- colnames(data_curated[IPQ_cols])
AbsPPM_columns <- colnames(data_curated[AbsPPM_cols])   # write col. number of Abs. ppm

pass_matrix <- c()
for (i in 1:nrow(data_curated)){
	# CREATE LIST THAT CONTAINS PASS_FAIL
	pass_list <- c()
	# THIS FOR LOOPS 'CHEATS' TO ITERATE OVER ALL ANALYTES
	for (j in 1:analytes_num) {
		pass = TRUE
	
		# IF ANALYTE FAIL SN, SET PASS TO FALSE
		if (is.na(data_curated[i,SN_columns[j]]) || data_curated[i,SN_columns[j]] < SN) {
			pass = FALSE
		}

		# IF ANALYTE FAILS IPQ, SET PASS TO FALSE
		if (is.na(data_curated[i,IPQ_columns[j]]) || data_curated[i,IPQ_columns[j]] > IPQ) {
			pass = FALSE
		}

		# IF ANALYTE FAILS PPM, SET PASS TO FALSE
		if (is.na(data_curated[i,AbsPPM_columns[j]]) || data_curated[i, AbsPPM_columns[j]] > PPM) {
			pass = FALSE
		}

		# IF PASS IS FALSE APPEND 0 TO PASS_LIST, ELSE 1
		if (pass == FALSE) {
			pass_list  <- c(pass_list, 0)
		 } else {
			pass_list  <- c(pass_list, 1)
		}
	}
	# JOIN PASS LISTS INTO PASS_MATRIX
	pass_matrix <- rbind(pass_matrix, pass_list)
	rownames(pass_matrix)[i] <- eval(parse(text=paste("data_curated$", meta_conCol, "[", i, "]", sep="")))
}
# ASSIGN COLUMN NAMES AND MERGE BOTH MATRICES
names_list <- c()
mass_list <- data.frame()
for (analyte in SN_columns) {
	parts = strsplit(analyte, "_")
	name = tail(parts[[1]], 1)
	mass_list = rbind(mass_list, c(name, parts[[1]][2]))
	name = paste("QC_",name,sep="")
	names_list = c(names_list, name)
}
colnames(pass_matrix) <- names_list
colnames(mass_list) <- c("Trait", "mass")
mass_list <- mass_list %>% mutate_at(c("mass"), as.numeric)
data_verdict <- cbind(data_curated, pass_matrix)
if(!is.null(colNameCohort)){
  pass_matrix <- cbind(data_verdict[[colNameCohort]], pass_matrix)
  colnames(pass_matrix)[1] <- "Cohort"
}
addWorksheet(wb, "analyte verdict")
writeData(wb, "analyte verdict", pass_matrix, rowNames = TRUE, keepNA = TRUE)
addWorksheet(wb, "masses")
writeData(wb, "masses", mass_list, rowNames = FALSE, keepNA = TRUE)
saveWorkbook(wb, file = outputFileName, overwrite = TRUE)
```

```{r Check quality controls}
# To be included in future
```

```{r Select analytes which passed criteria}
# select columns with 0/1 anaQC verdict
data_anaQC <- data_verdict %>% select(!!! rlang::syms(colNameSampleType), !!! rlang::syms(colNameCohort), colnames(data_verdict %>% select(matches("^QC_"))))

# Calculate (per biological group) %-age of spectra in which the analyte passed QC
table_percent <- data_anaQC %>%
                    group_by(!!! rlang::syms(colNameCohort), !!! rlang::syms(colNameSampleType)) %>%
                    summarise_all((list(percent = mean)))

writeData(wb, "analyte verdict", table_percent, startRow = nrow(data_verdict)+2, rowNames = FALSE, keepNA = TRUE)
## Generate pass list for all analytes per biological group
pass_matrix <- c()

for (i in 1:nrow(table_percent)){
  pass_list = c()
  for (j in colnames(table_percent[,-which(names(table_percent) %in% c(colNameCohort,colNameSampleType))])) {
		pass = TRUE
		if (table_percent[i,j] < k_Cutoff_per_group) {
			pass = FALSE
		}
		if (pass == FALSE) {
			pass_list  = c(pass_list, 0)
		} else {
			pass_list  = c(pass_list, 1)
		}
	}
	# JOIN PASS LISTS INTO PASS_MATRIX
	pass_matrix = rbind(pass_matrix, pass_list)
}
pass_matrix = data.frame(cbind(table_percent[,which(names(table_percent) %in% c(colNameCohort,colNameSampleType))], pass_matrix))

# MERGE BOTH MATRICES
colnames(pass_matrix) <- colnames(table_percent)
if(!is.null(colNameCohort)){
  rownames(pass_matrix) <- paste("pass_list_samplegroup_", pass_matrix[[which(names(table_percent) %in% c(colNameCohort))]], pass_matrix[[which(names(table_percent) %in% c(colNameSampleType))]], sep="")
} else {
  rownames(pass_matrix) <- paste("pass_list_samplegroup_", pass_matrix[[which(names(table_percent) %in% c(colNameSampleType))]], sep="")
}

colnames(pass_matrix) <- gsub("percent", "include", colnames(pass_matrix))
colnames(pass_matrix) <- gsub("QC_", "", colnames(pass_matrix))
writeData(wb, "analyte verdict", pass_matrix, startRow = nrow(data_verdict)+nrow(table_percent)+3, rowNames = FALSE, keepNA = TRUE)
if(!is.null(colNameCohort)){
  pass_matrix_sum <- pass_matrix %>% group_by(!!sym(colNameCohort)) %>% summarize(across(-c(1), max))
  pass_matrix_sum <- pass_matrix %>% 
                        select(!!sym(colNameCohort), !!sym(colNameSampleType)) %>% 
                        left_join(y=pass_matrix_sum, by=colNameCohort)
} else {
  pass_matrix_sum <- pass_matrix %>% filter(!(!!! rlang::syms(colNameSampleType) %in% c(typeNumTech, typeNumBlank))) %>% summarize(across(-c(1), max))
  pass_matrix_sum <- rbind(pass_matrix_sum, pass_matrix_sum[rep(1,nrow(pass_matrix %>% filter(!(!!! rlang::syms(colNameSampleType) %in% c(typeNumTech, typeNumBlank))))-1), ])
  pass_matrix_sum <- cbind(pass_matrix %>% filter(!(!!! rlang::syms(colNameSampleType) %in% c(typeNumTech, typeNumBlank))) %>% select(!!! rlang::syms(colNameSampleType)), pass_matrix_sum)
  pass_matrix_sum <- rbind(pass_matrix_sum, pass_matrix %>% filter(!!! rlang::syms(colNameSampleType) %in% c(typeNumTech, typeNumBlank)))
}
writeData(wb, "analyte verdict", pass_matrix_sum, startRow = nrow(data_verdict)+nrow(table_percent)+nrow(pass_matrix)+4, rowNames = FALSE, keepNA = TRUE)
saveWorkbook(wb, file = outputFileName, overwrite = TRUE)
# Number of analytes passed
rowSums(pass_matrix[,-which(names(table_percent) %in% c(colNameCohort,colNameSampleType))])-length(excludeTrait)
```

```{r Additional spectra quality control based on curated analytes}
totInt_curated <- rowSums(data_verdict %>% select(colnames(data_verdict %>% select(matches("^AbsArea-BG_")) %>% select(!matches(paste0('_', excludeTrait, '$', collapse = '|'))))) * data_verdict %>% select(colnames(data_verdict %>% select(matches("^QC_")) %>% select(!matches(paste0('_', excludeTrait, '$', collapse = '|')))))) # Exclude glycans mentioned in excludeTrait

data_verdict <- cbind(data_verdict, totInt_curated)

v_rowsums <- rowSums(data_verdict %>% select(colnames(data_verdict %>% select(matches("^QC_")) %>% select(!matches(paste0('_', excludeTrait, '$', collapse = '|'))))))

data_verdict <- cbind(data_verdict, v_rowsums)

##Calculate (per cohort and per biological group) maximum of analytes which passed QC and maximum of total intensity
data_pass <- data_verdict %>% 
        inner_join(y = data_verdict %>% 
              group_by(!!! rlang::syms(colNameCohort), !!! rlang::syms(colNameSampleType)) %>%
              summarise(max = max(v_rowsums), maxInt = max(totInt_curated), .groups="keep"), by = c(colNameCohort, colNameSampleType))

data_pass$pass <- ifelse(data_pass$v_rowsums < k_Cutoff_anal_num*data_pass$max | data_pass$totInt_curated < k_Cutoff_TotInt_cur*data_pass$maxInt, 0, 1)
data_pass <- data_pass %>% droplevels()
data_NOpass <- data_pass %>% filter(pass == 0) 
data_pass <- data_pass %>% filter(pass == 1) #select only spectra which passed the criteria

addWorksheet(wb, "curated_spectra_2")
writeData(wb, "curated_spectra_2", data_pass, rowNames = FALSE, keepNA = TRUE)
addWorksheet(wb, "removed_spectra_2")
writeData(wb, "removed_spectra_2", data_NOpass, rowNames = FALSE, keepNA = TRUE)
saveWorkbook(wb, file = outputFileName, overwrite = TRUE)
```

### Spectrum Quality Plots after second spectrum curation
```{r Plots after 2nd spectrum curation}
#Total Intensity of passed analytes vs.Number of passed analytes 
QCplot <- ggplot(data_pass, aes(x=data_pass[,"totInt_curated"], y=data_pass[,"v_rowsums"], color=!!sym(colNameSampleType))) +
            geom_point(shape=1) +
            scale_colour_manual(limits = levels(unique(data_pass[[colNameSampleType]])), values=figCols) +
            xlab("Total intensity") + ylab("Number of passed analytes")
QCplot+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                 panel.background = element_blank(), axis.line = element_line(colour = "black"))


#Boxplot
boxplot(data_pass[,"v_rowsums"], xlab="spectra kept", ylab="Number of passed analytes")
```

### Renormalization and output to excel
```{r Final selection}
### Renormalize relative intensities on the selected analytes
# ## ALTERNATIVE A: All groups together
#Prepare selected sub-dataset
data_pre_renorm <- data_pass %>% select(all_of(data_conCol), !!! rlang::syms(colNameCohort), !!! rlang::syms(colNameSampleType), colnames(data_pass %>% select(matches("^CorrArea-BG_")) %>% select(!matches(paste0('_', excludeTrait, '$', collapse = '|'))))) # indicate CorrArea-BG columns
cnames <- sapply(strsplit(colnames(data_pre_renorm), split='_', fixed=TRUE), function(x) (x[3]))
names(data_pre_renorm) <- cnames
names(data_pre_renorm)[which(is.na(cnames))] <- c(data_conCol, colNameCohort, colNameSampleType)

## Multiply data_pre_renorm with pass_matrix_sum: only glycans selected per group will be > 0.
# if(!is.null(colNameCohort)){
#   pass_matrix_sum <- data_pre_renorm %>% 
#                     select(all_of(data_conCol), !!sym(colNameCohort), !!sym(colNameSampleType)) %>% 
#                     left_join(y=pass_matrix_sum %>% select(!matches(paste0(excludeTrait, "_include"))) %>% select(-c(colNameSampleType)), by = c(colNameCohort))
# } else {
#   pass_matrix_sum <- data_pre_renorm %>% 
#                     select(all_of(data_conCol), !!sym(colNameCohort), !!sym(colNameSampleType)) %>% 
#                     left_join(y=pass_matrix_sum %>% select(!matches(paste0(excludeTrait, "_include"))), by = c(colNameSampleType))
# }
pass_matrix_sum <- data_pre_renorm %>% 
                    select(all_of(data_conCol), !!! rlang::syms(colNameCohort), !!! rlang::syms(colNameSampleType)) %>% 
                    left_join(y=pass_matrix_sum %>% select(!matches(paste0(excludeTrait, "_include"))), by = c(colNameCohort, colNameSampleType))

data_pre_renorm_select <- (data_pre_renorm %>%
                            select(-c(all_of(data_conCol), !!! rlang::syms(colNameCohort), !!! rlang::syms(colNameSampleType)))) * (pass_matrix_sum %>%
                            select(-c(all_of(data_conCol), !!! rlang::syms(colNameCohort), !!! rlang::syms(colNameSampleType))))

## ...and renormalize to row sum
data_pre_renorm_select <- cbind(data_pre_renorm_select, oldSum = rowSums(data_pre_renorm_select))
data_renorm <- data_pre_renorm_select %>% mutate(across(.cols = everything(), .fns = list("renorm" = ~./data_pre_renorm_select$oldSum))) %>% rename(newSum = oldSum_renorm)

# Merge renormalized data with nonGlyco information
renormalized = cbind(data_pre_renorm %>% select(all_of(data_conCol)), data_renorm %>% select(matches("_renorm$")))
colnames(renormalized) <- c(data_conCol, sapply(strsplit(colnames(renormalized[,2:ncol(renormalized)]), split='_', fixed=TRUE), function(x) (x[1])))
renormalized <- renormalized %>% select(colnames(renormalized[1]), colnames(renormalized[which(colSums(renormalized[2:ncol(renormalized)])!=0)+1]))
mass_list <- mass_list %>% filter(Trait %in% colnames(renormalized[,-which(colnames(renormalized) %in% c(all_of(data_conCol), colNameCohort, colNameSampleType))])) %>% arrange(., mass)
renormalized <- merge(metaData, renormalized)
addWorksheet(wb, "curated_masses")
writeData(wb, "curated_masses", mass_list, rowNames = FALSE, keepNA = TRUE)
addWorksheet(wb, "renormalized")
writeData(wb, "renormalized", renormalized, rowNames = FALSE, keepNA = TRUE)
saveWorkbook(wb, file = outputFileName, overwrite = TRUE)
```

### Additional analyte curation based on rSD
```{r Analyte curation based on rSD}
if(!is.null(RSD)){
  renormCV <- NULL
  renormData <- cbind(data_pre_renorm[1:2], data_renorm %>% select(matches("_renorm$"))) %>% group_by(!!sym(colNameSampleType))
  colnames(renormData) <- c(colnames(renormData)[1:2], sapply(strsplit(colnames(renormData[,3:ncol(renormData)]), split='_', fixed=TRUE), function(x) x[1]))
  renormMean <- renormData %>%
      summarise_at(vars(colnames(renormData[,-c(1,2)])), list(mean = mean))
  renormSD <-renormData %>%
      summarise_at(vars(colnames(renormData[,-c(1,2)])), list(sd = sd))
  renormCV <- cbind(renormSD[1],renormSD[-1]/renormMean[-1])
  colnames(renormCV) <- sapply(strsplit(colnames(renormCV), split='_', fixed=TRUE), function(x) x[1])
  colnamesCV <- colnames(renormCV %>% select(where(~is.numeric(.x) && any(.x < RSD/100))))
  selectCV <- renormData %>% select(xy_filename, !!sym(colNameSampleType), which(colnames(renormData) %in% colnamesCV))
  addWorksheet(wb, "renormalized_RSD_any")
  writeData(wb, "renormalized_RSD_any", selectCV, rowNames = FALSE, keepNA = TRUE)
  colnamesCV <- colnames(renormCV %>% select(where(~is.numeric(.x) && all(.x < RSD/100))))
  selectCV <- renormData %>% select(xy_filename, !!sym(colNameSampleType), which(colnames(renormData) %in% colnamesCV))
  addWorksheet(wb, "renormalized_RSD_all")
  writeData(wb, "renormalized_RSD_all", selectCV, rowNames = FALSE, keepNA = TRUE)
  saveWorkbook(wb, file = outputFileName, overwrite = TRUE)
} else {
 print("Not executed")
}
```

### Session Info
```{r}
sessionInfo()
```