---
title: "Glycan Data Check After Curation"
author: "Marco Bladergroen"
date: "version 4.0.0, 01-04-2025, executed `r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
editor_options:
  chunk_output_type: inline
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding,
  output_file = file.path(dirname(inputFile), paste0('Glycan Data Check After Curation ', format(Sys.time(), '%Y-%m-%d %H_%M_%S'), '.html'))) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=14, echo = TRUE)
```
# Version history:

Version 1.0 (Marco Bladergroen): date unknown
<ul></ul>
Version 3.1 (Marco Bladergroen): 2024-04-11
<ul></ul>
Version 3.1.1 (Marco Bladergroen): 2024-06-25
<ul></ul>
Version 4.0.0 (Marco Bladergroen): 2025-04-01
<ul>Script adapted to new data curation output (v4.0.1): New data curation output can produce zero-values. Upon selection and filtering this can create columns with all 0 or all NA values. Script adapted so that these values (columns) are not included. Technical check now performed per technical group (pool per cohort / Visucon).</ul>

# Libraries used
```{r libraries, include=FALSE}
library(tidyverse)
library(openxlsx)
library(ggplot2)
library(reshape2)
library(lubridate)
library(GMCM)
library(plotly)
library(data.table)
```
# Function Definitions
```{r functions}
lm_eqn <- function(m){
  eq <- if(coef(m)[2]<0){
          substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2, 
                    list(a = format(unname(coef(m)[1]), digits = 3),
                         b = format(unname(abs(coef(m)[2])), digits = 5),
                         r2 = format(summary(m)$r.squared, digits = 2)))
        } else {
          substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,
                    list(a = format(unname(coef(m)[1]), digits = 3),
                         b = format(unname(coef(m)[2]), digits = 5),
                         r2 = format(summary(m)$r.squared, digits = 2)))
        }
  as.character(as.expression(eq));
}
```
# Settings
```{r settings}
wd <- getwd()

### data file direct traits
traitFileName <- "<Name of your curated data file>.xlsx"
sheetNameTraits <- "renormalized"
traitConCol <- "xy_filename" # Type here the name of the column in the datafile connecting the data with the metadata. Also define when no metafile is used = unique identifier for sample
if(traitFileName!=""){
  traitFileName <- paste(wd, traitFileName, sep="/")
}

### data file derived traits
derivedFileName <- "<Name of the glycosylation traits file>.xlsx"
sheetNameDerived <- "derived_traits_curated"
derivedConCol <- "xy_filename" # Type here the name of the column in the datafile connecting the data with the metadata. Also define when no metafile is used = unique identifier for sample
if(derivedFileName!=""){
  derivedFileName <- paste(wd, derivedFileName, sep="/")
}

### meta file
metaFileName <- "<Name of your metaData file>.xlsx"
sheetNameMeta <- "metaData"
if(metaFileName!=""){
  metaFileName <- paste(wd, metaFileName, sep="/")
}
metaConCol <- "xy_filename" # Type here the name of the column in the metafile connecting the data with the metadata
nMetaCols <- NULL #11 number of metadata columns, set to NULL when to be determined by script

### IgG file
IgGFileName <- ""
sheetNameIgG <- ""

### technical file if different from data file
technicalCheck <- TRUE
technicalCheckFileName <- "" #2022-02-25_Derived_traits_MC_Batch_Corrected.xlsx
sheetNameTechnicalCheck <- "renormalized"

# Percentage cut-off from numeric to factorial or character to factorial
facToNum <- 0.05
facToChar <- 0.1

# Sample type number or names: you need a column SampleType in your metaData 
typeNumCase <- c(1,3,5)
typeNumControl <- c(2,4,6)
typeNumPool <- c(7,8,9)
typeNumVisucon <- 10

technicalColumnsToCheck <- c("SampleType", "PlateBarcode")
topN <- 25 # set top N items to check (highest peaks)

ageColumn <- "Age" #"Age"
sexColumn <- "Sex" #"Sex"

### include or exclude specific features
# use the features (values) of metadata to include in filter
# examples : 
#   - all variables should be included : NULL
#   - only male in agegroup 40-49 should be included : c("male", "agegroup") and list(c("1"), c("40-49"))(see also show_filter)
show_filter_include <- NULL # Set here which group you want to filter (either include or exclude)
include <- NULL #list(c("1"), c("1"))
show_filter_exclude <- NULL #c("SampleType")
exclude <- NULL #list(c(typeNumVisucon)) #list(c("BEAT6W"))
```

# Data import
```{r Read_data}
#Read traits if present
if(traitFileName!=""){
  if(sheetNameTraits != ""){
    traitData <- read.xlsx(traitFileName, sheet = sheetNameTraits, colNames = TRUE, sep=" ")
  } else {
    traitData <- read.xlsx(traitFileName, colNames = TRUE, sep=" ")
  }
}

#Read derived traits if present
if(derivedFileName!=""){
  if(sheetNameDerived != ""){
    derivedData <- read.xlsx(derivedFileName, sheet = sheetNameDerived, colNames = TRUE, sep=" ")
  } else {
    derivedData <- read.xlsx(derivedFileName, colNames = TRUE, sep=" ")
  }
}

# Read Metadata if present. If not, extract from traits or derived traits
if(metaFileName!="") {
  if(sheetNameMeta!=""){
    metaData <- read.xlsx(metaFileName, sheet=sheetNameMeta, colNames = TRUE, sep=" ", na.strings="")
  } else {
    metaData <- read.xlsx(metaFileName, colNames = TRUE, sep=" ", na.strings="")
  }
  metaData[, grepl("date", tolower(names(metaData)))] <- metaData[, grepl("date", tolower(names(metaData)))] %>% mutate(across(everything(), ~ replace(., is.na(.), 0))) %>% mutate(across(everything(), ~ as_date(., origin = "1899-12-30"))) %>% mutate(across(where(is.Date), ~ na_if(.x,as.Date("1899-12-30"))))
  metaData <- metaData %>% mutate(across(where(is.numeric), as.factor))
  metaData <- metaData %>% mutate(across(!where(lubridate::is.Date) & names(which(sapply(metaData,nlevels) > facToNum*nrow(metaData))), function(x) as.numeric((levels(x))[x])))
  metaData <- metaData %>% mutate(across(!where(is.numeric) & !where(lubridate::is.Date) & colnames(metaData[!colnames(metaData) %in% metaConCol]), as.factor))
  metaData <- metaData %>% mutate(across(!where(is.numeric) & !where(lubridate::is.Date) & names(which(sapply(metaData,nlevels) > facToChar*nrow(metaData))), as.character))
  nMetaCols <- ncol(metaData)
} else {
  if(is.null(nMetaCols)){
    if(traitFileName!=""){
      traitData[, grepl("date", tolower(names(traitData)))] <- traitData[, grepl("date", tolower(names(traitData)))] %>% mutate(across(everything(), ~ replace(., is.na(.), 0))) %>% mutate(across(everything(), ~ as_date(., origin = "1899-12-30"))) %>% mutate(across(where(is.Date), ~ na_if(.x,as.Date("1899-12-30"))))
      metaCols <- c(colnames(traitData %>% 
                               select(-names(which(sapply(traitData[,colSums(is.na(traitData %>% mutate(across(!where(is.numeric), as.numeric))))-colSums(is.na(traitData))==0] %>% mutate(across(where(is.numeric), as.factor)), nlevels) > facToChar*nrow(traitData)) & which(sapply(traitData, is.Date))))))
    } else {
      derivedData[, grepl("date", tolower(names(derivedData)))] <- derivedData[, grepl("date", tolower(names(derivedData)))] %>% mutate(across(everything(), ~ replace(., is.na(.), 0))) %>% mutate(across(everything(), ~ as_date(., origin = "1899-12-30"))) %>% mutate(across(where(is.Date), ~ na_if(.x,as.Date("1899-12-30"))))
      metaCols <- c(colnames(derivedData %>% select(-names(which(sapply(derivedData[,colSums(is.na(derivedData %>% mutate(across(!where(is.numeric), as.numeric))))-colSums(is.na(derivedData))==0] %>% mutate(across(where(is.numeric), as.factor)), nlevels) > facToChar*nrow(derivedData)) & which(sapply(derivedData, is.Date))))))
    }
    nMetaCols <- length(metaCols)
  } else {
    metaCols <- colnames(raw_data[1:nMetaCols])
  }
  if(traitFileName!=""){
    metaData <- traitData %>% select(1:all_of(nMetaCols)) %>% mutate(across(where(is.numeric), as.factor))
  } else {
    metaData <- derivedData %>% select(1:all_of(nMetaCols)) %>% mutate(across(where(is.numeric), as.factor))
  }
  metaData <- metaData %>% mutate(across(names(which(sapply(metaData,nlevels) > facToNum*nrow(metaData) & !sapply(metaData, is.Date))), function(x) as.numeric((levels(x))[x])))
  metaData <- metaData %>% mutate(across(!where(is.numeric) & metaCols[!metaCols %in% metaConCol], as.factor))
  metaData <- metaData %>% mutate(across(!where(is.numeric) & names(which(sapply(metaData,nlevels) > facToChar*nrow(metaData))), as.character))
  if(traitFileName!=""){
    metaConCol <- traitConCol
  } else {
    metaConCol <- derivedConCol
  }
}

#Make final table of trait data
if(traitFileName!=""){
  if(length(which(colnames(traitData) %in% colnames(metaData)[!colnames(metaData) %in% c(traitConCol)]))==0){
    allTraitData <- metaData %>% inner_join(y = traitData, by = c({metaConCol = traitConCol}))
  } else {
    allTraitData <- metaData %>% inner_join(y = traitData[,-which(colnames(traitData) %in% colnames(metaData)[!colnames(metaData) %in% c(traitConCol)])], by = c({metaConCol = traitConCol}))
  }
  nTraitCols <- ncol(allTraitData)-nMetaCols
}

#Make final table of derived trait data
if(derivedFileName!=""){
  if(length(which(colnames(derivedData) %in% colnames(metaData)[!colnames(metaData) %in% c(derivedConCol)]))==0){
    allDerivedData <- metaData %>% inner_join(y = derivedData, by = c({metaConCol = derivedConCol}))
  } else {
    allDerivedData <- metaData %>% inner_join(y = derivedData[,-which(colnames(derivedData) %in% colnames(metaData)[!colnames(metaData) %in% c(derivedConCol)])], by = c({metaConCol = derivedConCol}))
  }
  # filter out all columns with all NA or all 0 or 1
  allDerivedData.filter <- data.frame(Sums = colSums(allDerivedData[,(nMetaCols+1):ncol(allDerivedData)])!=0 & colSums(allDerivedData[,(nMetaCols+1):ncol(allDerivedData)])!=nrow(allDerivedData) & !is.na(colSums(allDerivedData[,(nMetaCols+1):ncol(allDerivedData)])))
  allDerivedData.filter <- rownames(allDerivedData.filter %>% filter(Sums == TRUE))
  allDerivedData <- cbind(allDerivedData[, 1:nMetaCols],allDerivedData[,allDerivedData.filter])
  nDerivedCols <- ncol(allDerivedData)-nMetaCols
}

#Make complete final table with all data combined
if(traitFileName!=""){
  allData <- allTraitData
} else {
  allData <- metaData
}
if((derivedFileName!="" && traitFileName=="")|(derivedFileName!="" && traitFileName!=derivedFileName)){
  if(length(which(colnames(allDerivedData) %in% colnames(metaData)[!colnames(metaData) %in% c(derivedConCol)]))==0){
    allData <- allData %>% inner_join(y = allDerivedData, by = c({metaConCol = derivedConCol}))
  } else {
    allData <- allData %>% inner_join(y = allDerivedData[,-which(colnames(allDerivedData) %in% colnames(metaData)[!colnames(metaData) %in% c(derivedConCol)])], by = c({metaConCol = derivedConCol}))
  }
}
allData <- allData %>% mutate_at(colnames(allData[,(ncol(metaData)+1):ncol(allData)]), as.numeric)
nDataCols <- ncol(allData)-nMetaCols

if(!is.null(include)){
  for(i in 1:length(show_filter_include)){
    allData <- allData %>% 
      filter(eval(parse(text = show_filter_include[[i]])) %in% include[[i]])
  }
}

if(!is.null(exclude)){
  for(i in 1:length(show_filter_exclude)){
    allData <- allData %>% 
      filter(!(eval(parse(text = show_filter_exclude[[i]])) %in% exclude[[i]]))
  }
}

#Import IgG data if present
if(IgGFileName !=""){
  IgGdata <- read.xlsx(IgGFileName, sheet = sheetNameIgG, colNames= TRUE)
  IgGdata <- IgGdata %>% filter(Identifier == "Total")
  IgGdata$Group = as.character(IgGdata$Group)

  combinedData <- left_join(traitData, IgGdata, by=c("sample_id" = "Group"))
  total_IgG <- combinedData %>% filter(Identifier == "Total") %>% group_by(beat_id) %>% filter(date_sampling == min(date_sampling))
}

# Extract data of technical samples (visucon and/or sample pool)
if(technicalCheckFileName != ""){
  if(sheetNameTechnicalCheck != ""){
    technicalData <- read.xlsx(technicalCheckFileName, sheet = sheetNameTechnicalCheck, colNames= TRUE)
  } else {
    technicalData <- read.xlsx(technicalCheckFileName, colNames= TRUE)
  }
} else {
  technicalData <- allData
}
technicalData <- technicalData %>% filter(SampleType %in% c(typeNumVisucon, typeNumPool))
technicalData <- technicalData[,which(technicalData %>% sapply(function(x)!all(is.na(x))))] %>% droplevels()
```

```{r technicalCheck per technical group}
if(technicalCheckFileName != "" | technicalCheck==TRUE && nrow(technicalData)>0){
  # do technical check first for traits then derived traits
  plts <- NULL
  techData<-list()
  for(i in 1:2){
    if(i==1){
      start_col <- which(str_detect(colnames(technicalData), "^[Hh]([0-9]+)[Nn]([0-9]+).*$") == TRUE)[1] # column number where the glycans start, starts with HxNy
      nMetaColsTech <- start_col-1
    } else {
      start_col <- which(str_detect(colnames(technicalData), "\\bTU\\b") == TRUE)[1] # column number where the derived traits start, starts with TU
    }
    
    if(!is.na(start_col)){ # Traits or derived traits found
      # First determine overall mean and sd for all compounds, grouped by sampletype
      if(i==1){
        overallMean <- technicalData %>%
                       group_by(SampleType) %>% 
                       select(all_of(c(which(colnames(technicalData) == "SampleType"), start_col:(start_col + nTraitCols - 1)))) %>% 
                       summarise(across(everything(),mean), .groups='drop') %>% 
                       as.data.frame() %>% 
                       pivot_longer(!SampleType,names_to = "Trait", values_to = "mean")
        overallSD <- technicalData %>%
                     group_by(SampleType) %>% 
                     select(all_of(c(which(colnames(technicalData) == "SampleType"), start_col:(start_col + nTraitCols - 1)))) %>% 
                     summarise(across(everything(),sd), .groups='drop') %>% 
                     as.data.frame() %>% 
                     pivot_longer(!SampleType,names_to = "Trait", values_to = "sd")
        overallCV <- overallMean %>% 
                     inner_join(y=overallSD, by=c("SampleType","Trait")) %>% 
                     mutate(CV=(sd/mean)*100) %>% 
                     replace(is.na(.), 0) %>% 
                     group_by(SampleType) %>% 
                     add_column(traitGroup = "Glycan", .before=1) %>%
                     inner_join(y=technicalData %>%
                                  group_by(SampleType) %>% 
                                  count(SampleType), by="SampleType")
      } else {
        overallMean <- technicalData %>%
                       group_by(SampleType) %>% 
                       select(all_of(c(which(colnames(technicalData) == "SampleType"), start_col:(start_col + nDerivedCols - 1))), -any_of(c("MM"))) %>% 
                       summarise(across(everything(),mean), .groups='drop') %>% 
                       as.data.frame() %>% 
                       pivot_longer(!SampleType,names_to = "Trait", values_to = "mean")
        overallSD <- technicalData %>%
                     group_by(SampleType) %>% 
                     select(all_of(c(which(colnames(technicalData) == "SampleType"), start_col:(start_col + nDerivedCols - 1))), -any_of(c("MM"))) %>% 
                     summarise(across(everything(),sd), .groups='drop') %>% 
                     as.data.frame() %>% 
                     pivot_longer(!SampleType,names_to = "Trait", values_to = "sd")
        overallCV <- rbind(overallCV, overallMean %>% 
                                      inner_join(y=overallSD, by=c("SampleType","Trait")) %>% 
                                      mutate(CV=(sd/mean)*100) %>% 
                                      replace(is.na(.), 0) %>% 
                                      add_column(traitGroup = "Trait", .before=1) %>%
                                      inner_join(y=technicalData %>%
                                                 group_by(SampleType) %>% 
                                                 count(SampleType), by="SampleType")) %>%
                     group_by(traitGroup, SampleType)
      }
    }
  }
  
  # First determine overall mean and sd for the first topX compounds per group
  overallTop <- overallCV %>% 
                arrange(desc(mean)) %>%
                slice_head(n=topN) %>%
                mutate(minCV=min(CV)) %>%
                mutate(maxCV=max(CV))
 
  cnames <- overallTop %>% 
            group_by(traitGroup, SampleType) %>% 
            select(traitGroup, SampleType, Trait) %>%
            mutate(Var=factor(paste0('Glycan_Trait_',row_number()),
                              levels = paste0('Glycan_Trait_',1:25),
                              labels = paste0('Glycan_Trait_',1:25),
                              ordered = T,
                              exclude = F)) %>% 
            complete(Var = Var) %>% pivot_wider(names_from = Var, values_from = Trait)

  for(n in 1:nrow(cnames)){
    techData[["data"]][[paste0(cnames$traitGroup[n],"_",cnames$SampleType[n])]] <- technicalData %>%
            filter(SampleType == cnames$SampleType[n]) %>%
            select(all_of(c(colnames(technicalData)[1:nMetaColsTech], as.character(cnames[n,3:(topN+2)]))))
  }

  for(tc in technicalColumnsToCheck){
    if(tc %in% colnames(technicalData))
    {
      techData[["Totals"]][[paste0("techTotalsGlycan",tc)]] <- list()
      techData[["Totals"]][[paste0("techTotalsTrait",tc)]] <- list()
      for(t in names(techData[["data"]])){
        techData[["summary"]][[paste0(t,"mean")]] <- data.frame(techData[["data"]][[t]]) %>%
          select(-any_of(c(colnames(technicalData)[1:nMetaColsTech])), any_of(c(tc))) %>% 
          relocate(any_of(c(tc)), .before=1) %>%
          group_by(!!! rlang::syms(tc)) %>% 
          summarise(across(everything(),mean), .groups='drop') %>% 
          as.data.frame() %>% 
          pivot_longer(!tc,names_to = "Trait", values_to = "mean")
        techData[["summary"]][[paste0(t,"sd")]] <- data.frame(techData[["data"]][[t]]) %>%
          select(-any_of(c(colnames(technicalData)[1:nMetaColsTech])), any_of(c(tc))) %>% 
          relocate(any_of(c(tc)), .before=1) %>%
          group_by(!!! rlang::syms(tc)) %>% 
          summarise(across(everything(),sd), .groups='drop') %>% 
          as.data.frame() %>% 
          pivot_longer(!tc,names_to = "Trait", values_to = "sd")
        techData[["summary"]][[paste0(t,"CV")]] <- data.frame(techData[["summary"]][[paste0(t,"mean")]]) %>% 
          inner_join(y=data.frame(techData[["summary"]][[paste0(t,"sd")]]), by = c(tc, "Trait")) %>% 
          mutate(CV=(sd/mean)*100) %>% 
          mutate(minCV=min(CV)) %>%
          mutate(maxCV=max(CV)) %>%
          replace(is.na(.), 0) %>% 
          mutate(SampleType = strsplit(t, "_")[[1]][2], .before=1) %>%
          mutate(traitGroup = strsplit(t, "_")[[1]][1], .before=1) %>%
          inner_join(y=techData[["data"]][[t]] %>%
                       group_by(!!sym(tc)) %>% 
                       count(!!sym(tc)), by = tc) %>%
          unite(Label, !!sym(tc), SampleType, sep="_", remove= FALSE) %>%
          mutate(SampleType = factor(SampleType), traitGroup=factor(traitGroup), Label = factor(Label)) %>%
          group_by(traitGroup, !!sym(tc))
        techData[["Totals"]][[paste0("techTotals",strsplit(t, "_")[[1]][1],tc)]] <- rbind(techData[["Totals"]][[paste0("techTotals",strsplit(t, "_")[[1]][1],tc)]], techData[["summary"]][[paste0(t,"CV")]])
      }
      
      for(plts in c(paste0("techTotalsGlycan",tc), paste0("techTotalsTrait",tc))){
        if(tc=="SampleType"){
          
        } else {
          
        }
        techData[["plots"]][[paste0("my_barplot_", substring(plts,11))]] <- data.frame(techData[["Totals"]][[plts]]) %>% 
          complete(nesting(!!!rlang::syms(unique(c(tc, "Label", "SampleType", "traitGroup")))), Trait) %>%
          group_by(!!sym(tc)) %>%
          ggplot(aes(x=Trait, y=mean, fill=Label)) +
          geom_col(width=0.7, position=position_dodge(0.7), na.rm = FALSE) +
          geom_errorbar(aes(x=Trait, ymin=mean-sd, ymax=mean+sd), linewidth=0.15, width=0.3, alpha=0.9, position=position_dodge(.7)) +
          ggtitle("Variation") +
          theme(axis.text.x = element_text(angle = 45,
                                           hjust = 1,
                                           vjust = 1)) +
          labs(x=plts, y="mean(relative abundance)")
        figFilename <- paste0("Repeatability ", substring(plts,11), " ", format(Sys.time(), "%Y-%m-%d %H_%M_%OS2"), ".pdf", sep="")
        ggsave(file=figFilename, plot=techData[["plots"]][[paste0("my_barplot_", substring(plts,11))]], width=10, height=7, limitsize = FALSE)
      }
    }
  }
  
  print("Summary overall")
  print(overallCV %>% filter(!(CV==0)))
  
  print(paste0("Summary for the top-",topN," (highest mean) traits per technical group"))
  print(overallTop)
  
  print("CV averages per technical group")
  print(overallCV %>% select(traitGroup, SampleType, CV) %>% filter(!(CV==0)) %>% summarize(averageCV = mean(CV), minCV = min(CV), maxCV = max(CV), n = n()))
  
  print(paste0("CV averages per technical group for the top-",topN," (highest mean)"))
  print(overallTop %>% select(traitGroup, SampleType, CV) %>% summarize(averageCV = mean(CV), minCV = min(CV), maxCV = max(CV)))
}
```

```{r technicalCheck per sample group}
if(technicalCheck==TRUE){
  # calculated CV for cases and controls: first for traits then derived traits
  plts <- NULL
  caseControlData<-list()
  for(i in 1:2){
    if(i==1){
      start_col <- which(str_detect(colnames(allData), "^[Hh]([0-9]+)[Nn]([0-9]+).*$") == TRUE)[1] # column number where the glycans start, starts with HxNy
      #nMetaColsTech <- start_col-1
    } else {
      start_col <- which(str_detect(colnames(allData), "\\bTU\\b") == TRUE)[1] # column number where the derived traits start, starts with TU
    }
    
    if(!is.na(start_col)){ # Traits or derived traits found
      # First determine overall mean and sd for all compounds, grouped by sampletype
      if(i==1){
        overallMean <- allData %>%
                       filter(SampleType %in% c(typeNumCase, typeNumControl)) %>%
                       group_by(SampleType) %>% 
                       select(all_of(c(which(colnames(allData) == "SampleType"), start_col:(start_col + nTraitCols - 1)))) %>% 
                       summarise(across(everything(),mean), .groups='drop') %>% 
                       as.data.frame() %>% 
                       pivot_longer(!SampleType,names_to = "Trait", values_to = "mean")
        overallSD <- allData %>%
                     filter(SampleType %in% c(typeNumCase, typeNumControl)) %>%
                     group_by(SampleType) %>% 
                     select(all_of(c(which(colnames(allData) == "SampleType"), start_col:(start_col + nTraitCols - 1)))) %>% 
                     summarise(across(everything(),sd), .groups='drop') %>% 
                     as.data.frame() %>% 
                     pivot_longer(!SampleType,names_to = "Trait", values_to = "sd")
        overallCV <- overallMean %>% 
                     inner_join(y=overallSD, by=c("SampleType","Trait")) %>% 
                     mutate(CV=(sd/mean)*100) %>% 
                     replace(is.na(.), 0) %>% 
                     group_by(SampleType) %>% 
                     add_column(traitGroup = "Glycan", .before=1) %>%
                     inner_join(y=allData %>%
                                  group_by(SampleType) %>% 
                                  count(SampleType), by="SampleType")
      } else {
        overallMean <- allData %>%
                       filter(SampleType %in% c(typeNumCase, typeNumControl)) %>%
                       group_by(SampleType) %>% 
                       select(all_of(c(which(colnames(allData) == "SampleType"), start_col:(start_col + nDerivedCols - 1))), -any_of(c("MM"))) %>% 
                       summarise(across(everything(),mean), .groups='drop') %>% 
                       as.data.frame() %>% 
                       pivot_longer(!SampleType,names_to = "Trait", values_to = "mean")
        overallSD <- allData %>%
                     filter(SampleType %in% c(typeNumCase, typeNumControl)) %>%
                     group_by(SampleType) %>% 
                     select(all_of(c(which(colnames(allData) == "SampleType"), start_col:(start_col + nDerivedCols - 1))), -any_of(c("MM"))) %>% 
                     summarise(across(everything(),sd), .groups='drop') %>% 
                     as.data.frame() %>% 
                     pivot_longer(!SampleType,names_to = "Trait", values_to = "sd")
        overallCV <- rbind(overallCV, overallMean %>% 
                                      inner_join(y=overallSD, by=c("SampleType","Trait")) %>% 
                                      mutate(CV=(sd/mean)*100) %>% 
                                      replace(is.na(.), 0) %>% 
                                      add_column(traitGroup = "Trait", .before=1) %>%
                                      inner_join(y=allData %>%
                                                 group_by(SampleType) %>% 
                                                 count(SampleType), by="SampleType")) %>%
                     group_by(traitGroup, SampleType)
      }
    }
  }
  
  # Determine overall mean and sd for the first topX compounds per group
  overallTop <- overallCV %>% 
                arrange(desc(mean)) %>%
                slice_head(n=topN) %>%
                mutate(minCV=min(CV)) %>%
                mutate(maxCV=max(CV))
 
  cnames <- overallTop %>% 
            group_by(traitGroup, SampleType) %>% 
            select(traitGroup, SampleType, Trait) %>%
            mutate(Var=factor(paste0('Glycan_Trait_',row_number()),
                              levels = paste0('Glycan_Trait_',1:25),
                              labels = paste0('Glycan_Trait_',1:25),
                              ordered = T,
                              exclude = F)) %>% 
            complete(Var = Var) %>% pivot_wider(names_from = Var, values_from = Trait)
  
  for(n in 1:nrow(cnames)){
    caseControlData[["data"]][[paste0(cnames$traitGroup[n],"_",cnames$SampleType[n])]] <- allData %>%
            filter(SampleType == cnames$SampleType[n]) %>%
            select(all_of(c(colnames(technicalData)[1:nMetaColsTech], as.character(cnames[n,3:(topN+2)]))))
  }
  
  for(tc in technicalColumnsToCheck){
    if(tc %in% colnames(technicalData))
    {
      caseControlData[["Totals"]][[paste0("caseControlTotalsGlycan",tc)]] <- list()
      caseControlData[["Totals"]][[paste0("caseControlTotalsTrait",tc)]] <- list()
      for(t in names(caseControlData[["data"]])){
        caseControlData[["summary"]][[paste0(t,"mean")]] <- data.frame(caseControlData[["data"]][[t]]) %>%
          select(-any_of(c(colnames(technicalData)[1:nMetaColsTech])), any_of(c(tc))) %>% 
          relocate(any_of(c(tc)), .before=1) %>%
          group_by(!!! rlang::syms(tc)) %>%  
          summarise(across(everything(),mean), .groups='drop') %>% 
          as.data.frame() %>% 
          pivot_longer(!tc,names_to = "Trait", values_to = "mean")
        caseControlData[["summary"]][[paste0(t,"sd")]] <- data.frame(caseControlData[["data"]][[t]]) %>%
          select(-any_of(c(colnames(technicalData)[1:nMetaColsTech])), any_of(c(tc))) %>% 
          relocate(any_of(c(tc)), .before=1) %>%
          group_by(!!! rlang::syms(tc)) %>% 
          summarise(across(everything(),sd), .groups='drop') %>% 
          as.data.frame() %>% 
          pivot_longer(!tc,names_to = "Trait", values_to = "sd")
        caseControlData[["summary"]][[paste0(t,"CV")]] <- data.frame(caseControlData[["summary"]][[paste0(t,"mean")]]) %>% 
          inner_join(y=data.frame(caseControlData[["summary"]][[paste0(t,"sd")]]), by = c(tc, "Trait")) %>% 
          mutate(CV=(sd/mean)*100) %>% 
          mutate(minCV=min(CV)) %>%
          mutate(maxCV=max(CV)) %>%
          replace(is.na(.), 0) %>% 
          mutate(SampleType = strsplit(t, "_")[[1]][2], .before=1) %>%
          mutate(traitGroup = strsplit(t, "_")[[1]][1], .before=1) %>%
          inner_join(y=caseControlData[["data"]][[t]] %>%
                       group_by(!!sym(tc)) %>% 
                       count(!!sym(tc)), by = tc) %>%
          unite(Label, !!sym(tc), SampleType, sep="_", remove= FALSE) %>%
          mutate(SampleType = factor(SampleType), traitGroup=factor(traitGroup), Label = factor(Label)) %>%
          group_by(traitGroup, !!sym(tc))
        caseControlData[["Totals"]][[paste0("caseControlTotals",strsplit(t, "_")[[1]][1],tc)]] <- rbind(caseControlData[["Totals"]][[paste0("caseControlTotals",strsplit(t, "_")[[1]][1],tc)]], caseControlData[["summary"]][[paste0(t,"CV")]])
      }
      
      for(plts in c(paste0("caseControlTotalsGlycan",tc), paste0("caseControlTotalsTrait",tc))){
        caseControlData[["plots"]][[paste0("my_barplot_", substring(plts,18))]] <- data.frame(caseControlData[["Totals"]][[plts]]) %>% 
          complete(nesting(!!!rlang::syms(unique(c(tc, "Label", "SampleType", "traitGroup")))), Trait) %>%
          group_by(!!sym(tc)) %>%
          ggplot(aes(x=Trait, y=mean, fill=Label)) +
          geom_col(width=0.7, position=position_dodge(0.7), na.rm = FALSE) +
          geom_errorbar(aes(x=Trait, ymin=mean-sd, ymax=mean+sd), linewidth=0.15, width=0.3, alpha=0.9, position=position_dodge(.7)) +
          ggtitle("Variation") +
          theme(axis.text.x = element_text(angle = 45,
                                           hjust = 1,
                                           vjust = 1)) +
          labs(x=plts, y="mean(relative abundance)")
        figFilename <- paste0("Variation ", substring(plts,18), " ", format(Sys.time(), "%Y-%m-%d %H_%M_%OS2"), ".pdf", sep="")
        ggsave(file=figFilename, plot=caseControlData[["plots"]][[paste0("my_barplot_", substring(plts,18))]], width=10, height=7, limitsize = FALSE)
      }
    }
  }

  print("Summary overall")
  print(overallCV %>% filter(!(CV==0)))
  
  print(paste0("Summary for the top-",topN," (highest mean) traits per biological group"))
  print(overallTop)
  
  print("CV averages per biological group")
  print(overallCV %>% select(traitGroup, SampleType, CV) %>% filter(!(CV==0)) %>% summarize(averageCV = mean(CV), minCV = min(CV), maxCV = max(CV), n = n()))
  
  print(paste0("CV averages per biological group for the top-",topN," (highest mean)"))
  print(overallTop %>% select(traitGroup, SampleType, CV) %>% summarize(averageCV = mean(CV), minCV = min(CV), maxCV = max(CV)))
}
```

# Known Age trends
### H3N4F1 and H3N5F1 up, H4N4F1 zero, H5N4F1 and A2FS0G down

```{r Age_dependencies, out.width ="100%", fig.height =15}
#if(!isTRUE(all.equal(allData, metaData, convert = TRUE))){
  # Paper Gerda: https://onlinelibrary.wiley.com/doi/10.1002/cam4.3439
  # Non published data from Viktoria
  if(!is.null(ageColumn)){
  ageCheck <- c("H3N4F1", "H3N4F1E0L0", "H3N5F1", "H3N5F1E0L0", "H4N4F1", "H4N4F1E0L0", "H5N4F1", "H5N4F1E0L0", "A2FS0G") # should go (up, up, equal, down, down)

  ageColumn = ensym(ageColumn)
  ageData <- allData[,c(1:nMetaCols, which(colnames(allData) %in% ageCheck))]
#  ageData <- allData %>% select(which(colnames(allData) %in% colnames(metaData), which(colnames(allData) %in% ageCheck)))
  # scale data (minus mean and divide by SD)
    # scale data (minus mean and divide by SD)
  ageData.scaled <- cbind(allData[,c(1:nMetaCols)], t((t(ageData[,which(colnames(ageData) %in% ageCheck)]) - colMeans(ageData[,which(colnames(ageData) %in% ageCheck)])) / GMCM:::colSds(as.matrix(ageData[,which(colnames(ageData) %in% ageCheck)])))) %>% filter(!is.na({{ ageColumn }}))
  for(gly in colnames(ageData.scaled[,which(colnames(ageData.scaled) %in% ageCheck)])) {
    ageData.scaled <- eval(parse(text=paste("cbind(ageData.scaled, ", gly, "resid = resid(lm( formula = ", gly, " ~ ", ageColumn, ", data=ageData.scaled)))", sep="")))
  }
  ageData <- ageData.scaled %>% filter(!(SampleType %in% c(typeNumVisucon, typeNumPool))) %>% select(c(1:all_of(nMetaCols), which(colnames(ageData.scaled) %in% ageCheck))) %>% reshape2::melt(measure = which(colnames(ageData.scaled) %in% ageCheck), varnames=names(dimnames(ageData.scaled)), value.name = "value")
  ageData.resid <- ageData.scaled %>% filter(!(SampleType %in% c(typeNumVisucon, typeNumPool))) %>% select(c(1:all_of(nMetaCols), which(colnames(ageData.scaled) %in% paste(ageCheck, "resid", sep=""))))
  ageData.resid <- ageData.resid %>% reshape2::melt(measure = which(colnames(ageData.resid) %in% paste(ageCheck, "resid", sep="")), varnames=names(dimnames(ageData.resid)), value.name = "value")
  nTraits <- length(which(colnames(ageData.scaled) %in% ageCheck))
  nCols <- if(nTraits==1) { 1 }else{ 2 }
  
  if(is.factor(ageData[[ageColumn]])){
    ageData <- ageData %>% mutate_at(c(paste(ageColumn)), function(x) as.numeric((levels(x))[x]))
  }
  
  minmax <- cbind((ageData %>%
    dplyr::group_by(variable) %>%
    dplyr::summarise(minx = min({{ ageColumn }}), maxy = max(value))), (ageData %>% 
    dplyr::group_by(variable) %>%
    do(label = lm_eqn(lm( formula = eval(parse(text = paste("value ~ ", ageColumn, sep=""))), data = .))))[,2], (ageData %>% dplyr::group_by(variable) %>% do(model=lm( formula = eval(parse(text = paste("value ~ ", ageColumn, sep=""))), data = .)))[2])
  
  if(is.factor(ageData.resid[[ageColumn]])){
    ageData.resid <- ageData.resid %>% mutate_at(c(paste(ageColumn)), function(x) as.numeric((levels(x))[x]))
  }
  minmax.resid <- cbind((ageData.resid %>% 
    dplyr::group_by(variable) %>%
    dplyr::summarise(minx = min({{ ageColumn }}), maxy = max(value))), (ageData.resid %>% 
    dplyr::group_by(variable) %>%
    do(label = lm_eqn(lm( formula = eval(parse(text = paste("value ~ ", ageColumn, sep=""))), data = .))))[,2], (ageData.resid %>% dplyr::group_by(variable) %>% do(model=lm( formula = eval(parse(text = paste("value ~ ", ageColumn, sep=""))), data = .)))[2])
  agePlot <- ageData %>%
      ggplot(aes(x = {{ ageColumn }},
                 y = value)) + geom_point(aes(colour = SampleType)) + 
      geom_smooth(method="lm", formula = y ~ x)  +
      facet_wrap( ~ variable, scales="free", ncol = nCols) +
      geom_text(data = minmax, x = minmax$minx, y = minmax$maxy, label = minmax$label,
             hjust = 0, parse = TRUE)
  if(is.factor(ageData[[ageColumn]])){
    agePlot <- agePlot + geom_boxplot(aes(fill = {{ageColumn}}), outlier.shape = NA)
  }
  modelhist <- ageData.resid %>%
      ggplot(aes(x = value)) +
      geom_histogram(binwidth = 0.05) +
      facet_wrap( ~ variable, scales="free", ncol = nCols)
  modelQQ <- ageData.resid %>%
      ggplot(aes(sample = value)) +
      geom_qq() +
      geom_qq_line() +
      facet_wrap( ~ variable, scales="free", ncol = nCols)
  homeoplot <- ageData.resid %>%
      ggplot(aes(x = {{ ageColumn }},
                 y = value)) + geom_point(aes(colour = SampleType)) + 
      geom_smooth(method="lm", formula = y ~ x) +
      geom_text(data = minmax.resid, x = minmax.resid$minx, y = minmax.resid$maxy, label = minmax.resid$label,
             hjust = 0, parse = TRUE) +
      facet_wrap( ~ variable, scales="free", ncol = nCols)

  if(length(which(colnames(allData) %in% ageCheck))>0){
    plot(agePlot)
    plot(modelhist)
    plot(modelQQ)
    plot(homeoplot)
  }
} else {
  print("Not executed")
}
```

# Known Sex trends
### Male should be higher

```{r Sex_dependencies, out.width ="100%", out.height ="100%"}
#if(!isTRUE(all.equal(allData, metaData, convert = TRUE)) && derivedFileName!=""){
sexCheck <- c("A3F", "A4F") #Especially derived traits: should be higher in male
if(!is.null(sexColumn) & derivedFileName!="" & length(which(colnames(allData) %in% sexCheck))>1){
  # Paper Gerda: https://onlinelibrary.wiley.com/doi/10.1002/cam4.3439
  # Non published data from Viktoria
  sexData <- allData[,c(1:nMetaCols, which(colnames(allData) %in% sexCheck))]
  sexData <- sexData %>% filter(!(SampleType %in% c(typeNumVisucon, typeNumPool))) %>% reshape2::melt(measure = which(colnames(sexData) %in% sexCheck), varnames=names(dimnames(sexData)), value.name = "value")
  sexColumn = ensym(sexColumn)
  sexPlot <- sexData %>%
      ggplot(aes(x = {{ sexColumn }},
                 y = value)) +
      geom_boxplot(aes(fill = {{sexColumn}}), outlier.shape = NA) +
      geom_point(aes(colour = SampleType)) + 
      facet_wrap( ~ variable, scales="free", ncol = 2)
  if(length(which(colnames(allData) %in% sexCheck))>0){
    plot(sexPlot)
  }
} else {
  print("Not executed: data (glycosylation traits A3F and A4F) not available")
}
```

# Total IgG galactosylation vs. A2G

```{r}
if(traitFileName != "" && IgGFileName !=""){
  ggscatter(total_IgG, x = "IgG1.galactosylation", y = "A2G", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Total IgG galactosylation", ylab = "A2G")

  #If you want, you can easily color the any variable (e.g. ICU, non-ICU) by adding to the above code.

  # ggscatter(data, x = "total_IgG_galactosylation", y = "A2G",
  #           add = "reg.line", conf.int = TRUE,
  #           cor.coef = TRUE, cor.method = "spearman",
  #           xlab = "Total IgG galactosylation", ylab = "A2G") +
  # stat_cor(aes(color = ward))
} else {
  print("Not executed")
}
```
# Total IgG galactosylation vs. A2FS0G

```{r}
if(traitFileName != "" && IgGFileName !=""){
  ggscatter(total_IgG, x = "IgG1.galactosylation", y = "A2FS0G", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Total IgG galactosylation", ylab = "A2FS0G")
} else {
  print("Not executed")
}
```

# Total IgG galactosylation vs. CG

```{r}
if(traitFileName != "" && IgGFileName !=""){
  ggscatter(total_IgG, x = "IgG1.galactosylation", y = "CG", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Total IgG galactosylation", ylab = "CG")
} else {
  print("Not executed")
}
```

# Total IgG galactosylation vs. TA2FS0

```{r}
if(traitFileName != "" && IgGFileName !=""){
  ggscatter(total_IgG, x = "IgG1.galactosylation", y = "TA2FS0", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Total IgG galactosylation", ylab = "TA2FS0")
} else {
  print("Not executed")
}
```

# IgG1 bisection vs. CB

```{r}
if(traitFileName != "" && IgGFileName !=""){
  ggscatter(total_IgG, x = "IgG1.bisection", y = "CB", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "IgG1 bisection", ylab = "CB")
} else {
  print("Not executed")
}
```

# IgG1 bisection vs. A2B

```{r}
if(traitFileName != "" && IgGFileName !=""){
  ggscatter(total_IgG, x = "IgG1.bisection", y = "A2B", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "IgG1 bisection", ylab = "A2B")
} else {
  print("Not executed")
}
```